---
title: "Diagnosen"
author: "Saverio Fontana"
date: "2025-03-05"
output: html_document
---

```{r setup, include=FALSE}
# define paths
# for data import
import_path <- "02_data/01_raw_data/"
# for export of inspection objects (spreadsheets)
inspect_path <- "03_spreadsheets/"
# for export of clean data
clean_path <- "02_data/02_clean_data/"

knitr::opts_chunk$set(echo = TRUE)

library(data.table)
library(rio)
library(dplyr)
library(tidyr)

```

```{r Diagnosen}
# function for data import from .csv files
import_dataset <- function(...) {
  import_list(..., rbind = TRUE, setclass = "data.table")
}

# import Diagnosen data set
Diagnosen <- import_dataset(paste0(import_path, 
                                "[SDS].[v_Export_Diagnosen_tra_sds].csv"))

#Diagnosen is an Excel w/ many sheets. I will import them all 
all_data <- import_list(paste0(import_path, "Diagnosen.xlsx"), setclass = "data.table")

# Let's see sheet names
names(all_data)

# Access a specific sheet
Diagnosen2 <- all_data[["(SDS).(v_Export_Diagnosen_tra_s"]]

```

The other sheets in the Excel seems more like summary and counting table (e.g. Pivog Table) than useful data, so Iwill stick for the analysis w/ the first sheet. Maybe we can discuss about the others.

```{r Comparing datasets}
data1 <- Diagnosen
data2 <- Diagnosen2

# Identify common cols
common_cols <- intersect(names(data1), names(data2))
extra_cols <- setdiff(names(data2), names(data1))

# Order cols (just in case)
data2_ord <- data2[, c(common_cols, extra_cols), with = FALSE]


# Key identifier
key_cols <- c("research_case_id", "Name", "Code")

# Order using key
data1 <- data1 %>% arrange(across(all_of(key_cols)))
data2 <- data2 %>% arrange(across(all_of(key_cols)))

dim(data1)
dim(data2)

#same nmbr of rows, but different columns 


# Compare different rows
missing_in_data2 <- anti_join(data1, data2, by = key_cols)
missing_in_data1 <- anti_join(data2, data1, by = key_cols)

dim(missing_in_data2)
dim(missing_in_data1)
```

Upon inspection, it's clear the same rows are different. The only thing that changes is the formatting, cause there is a blank space at the end of the column name of missing_in_data2.

we can then conclude that it is the same dataframe. For this reason we will work with Diagnosen2, which has more variables. 


```{r cleaning1}
# Create a translation dictionary
translation_dict <- c(
  "Code" = "Code", 
  "Dat" = "Date", 
  "Displayname" = "Display Name", 
  "Kat" = "Category", 
  "Name" = "Name", 
  "research_case_id" = "Research Case ID", 
  "research_id" = "Research ID", 
  "Version" = "Version", 
  "Blutung" = "Bleeding", 
  "Fraktur" = "Fracture", 
  "Subduralblutung" = "Subdural Bleeding", 
  "Epiduralblutung" = "Epidural Bleeding", 
  "Subarachnoidalblutung" = "Subarachnoid Bleeding", 
  "Kontusion" = "Contusion", 
  "Diffuse Hirnverletzungen und Riss/Shearing" = "Diffuse Brain Injuries and Ripping/Shearing", 
  "Andere Blutungen" = "Other Bleeding", 
  "Schädelbasisfraktur" = "Skull Base Fracture", 
  "Schädelkalottenfaktur" = "Skull Cap Fracture", 
  "Bewusstlosigkeit" = "Unconsciousness", 
  "Gehrinerschütterung" = "Concussion", 
  "Hirnödem" = "Brain Edema", 
  "Compressio cerebri" = "Brain Compression", 
  "Gehirnerschütterung Korrekturformel" = "Concussion Correction Formula", 
  "Gehirnerschütterung Korrektur inkl. Bewusstlosigkeit bei SHT" = "Concussion Correction including Unconsciousness in TBI"
)

# Overwrite the column names using the translation dictionary
colnames(Diagnosen2) <- translation_dict[colnames(Diagnosen2)]

# Convert column names to lowercase and replace spaces with underscores
names(Diagnosen2) <- tolower(gsub(" ", "_", names(Diagnosen2)))

# Rename the column to remove special characters (replace / with _)
names(Diagnosen2) <- gsub("/", "_", names(Diagnosen2))

# Convert "Date" column to Date type
Diagnosen2$date <- as.Date(Diagnosen2$date, format = "%Y-%m-%d %H:%M:%S")


#Remove some columns: 
#The "code" column i the ICD-10 (cause the "category" column has only one value: ICD10). I removed both columns, if noone is contrary....
Diagnosen2 <- Diagnosen2 %>%
  select(-code, -category)




unique(Diagnosen2$version)
#the column "version" contains various version of this ICD. I decided to remove also that column
#Also, consistent with the other script, I only keep the case identifier, not the other one :
Diagnosen2 <- Diagnosen2 %>%
  select(-version, -research_id)
```





```{r cleaning2}

#check missing values
table(is.na(Diagnosen2))


# Count NAs per column
na_count_per_column <- colSums(is.na(Diagnosen2))
na_count_per_column[na_count_per_column > 0]  # Display columns with NAs


# Calculate percentage of NA values per column
na_percentage <- colSums(is.na(Diagnosen2)) / nrow(Diagnosen2) * 100
na_percentage[na_percentage > 0]  # Display columns with NAs and their percentage

#the percentage is sooo low. We can inspect these rows. 

na_any_column_df <- Diagnosen2[!complete.cases(Diagnosen2), ]
#If we want we can keep them cause the missing is only from Date and a super small part ( 0.02%) from display_name. 

#I will remove them for now, I dont think it has a big impact....

#Anyway since some "display_name" miss, we could work with "name" which has 0 missing values. We can discuss it. 


# Remove rows with NAs
Diagn_clean <- Diagnosen2[complete.cases(Diagnosen2), ]


# remove duplicate row
Diagn_clean <- unique(Diagn_clean)

# View the cleaned dataframe
head(Diagn_clean)
```

After this cleaning, I looked it up for an automatic translator ( for the column "name"). There are some libraries if we want, but it's a bit of work cause we should setup an API for google translate (not that hard tbh, but we can discuss)

```{r Work on text}
# Clean up the text data
Diagn_clean$display_name <- tolower(Diagn_clean$display_name)  # Convert to lowercase
Diagn_clean$display_name <- gsub("\\s+", " ", Diagn_clean$display_name)  # Remove extra spaces
Diagn_clean$display_name <- trimws(Diagn_clean$display_name)  # Remove leading and trailing spaces

Diagn_clean$name <- tolower(Diagn_clean$name)  # Convert to lowercase
Diagn_clean$name <- gsub("\\s+", " ", Diagn_clean$name)  # Remove extra spaces
Diagn_clean$name <- trimws(Diagn_clean$name)  # Remove leading and trailing spaces


# Frequency plot for display_name
library(ggplot2)
top_display_names <- as.data.frame(table(Diagn_clean$display_name))
colnames(top_display_names) <- c("display_name", "count")
top_display_names <- top_display_names[order(-top_display_names$count), ]
head(top_display_names, 10)  # Show top 10

ggplot(top_display_names[1:10, ], aes(x = reorder(display_name, -count), y = count)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(title = "Top 10 Most Common Display Names", x = "Display Name", y = "Count")
```

```{r Handling IDs}
# Check for duplicate patient IDs
dupes <- Diagn_clean %>%
  group_by(research_case_id) %>%
  filter(n() > 1)
print(paste("Number of duplicate patient entries:", nrow(dupes)))
```

Upon inspection, each case can have multiple rows cause I suppose for each injury they fill out a new row w/ a new description of the injury. 

This lead to the question: how do we handle all these injuries? Do we use the column "name" or "display_name"? 

```{r display_name vs name}
names <- Diagn_clean %>% filter(display_name != name) %>% select(display_name, name) %>% unique()
names[1:10,]
```

They honestly seem super similar.... but this is up to you guys (German is too much)
I will stick with display_name as Vale did, we can always change this choice.


```{r}
#Drop variable name
Diagn_clean <- Diagn_clean %>%
  select(-name)


#Aggregates each patient to have a better df
Diagnosis_cleaned <- Diagn_clean %>%
  group_by(research_case_id) %>%
  summarize(
    date = first(date),  # Keep the first date for each individual
    bleeding = max(bleeding == 1, na.rm = TRUE),
    fracture = max(fracture == 1, na.rm = TRUE),
    subdural_bleeding = max(subdural_bleeding == 1, na.rm = TRUE),
    epidural_bleeding = max(epidural_bleeding == 1, na.rm = TRUE),
    subarachnoid_bleeding = max(subarachnoid_bleeding == 1, na.rm = TRUE),
    contusion = max(contusion == 1, na.rm = TRUE),
    diffuse_brain_injuries_and_ripping_shearing = max(diffuse_brain_injuries_and_ripping_shearing == 1, na.rm = TRUE),
    other_bleeding = max(other_bleeding == 1, na.rm = TRUE),
    skull_base_fracture = max(skull_base_fracture == 1, na.rm = TRUE),
    skull_cap_fracture = max(skull_cap_fracture == 1, na.rm = TRUE),
    unconsciousness = max(unconsciousness == 1, na.rm = TRUE),
    concussion = max(concussion == 1, na.rm = TRUE),
    brain_edema = max(brain_edema == 1, na.rm = TRUE),
    brain_compression = max(brain_compression == 1, na.rm = TRUE),
    concussion_correction_formula = max(concussion_correction_formula == 1, na.rm = TRUE),
    concussion_correction_including_unconsciousness_in_tbi = max(concussion_correction_including_unconsciousness_in_tbi == 1, na.rm = TRUE),
    injury_descriptions = paste(unique(display_name), collapse = ", "),  # Concatenate unique injury descriptions
    num_injuries = n_distinct(display_name)  # Count distinct injury descriptions (new col)
  )


# Convert binary variables to factors (we could also to logical if we wwant)
Diagnosis_cleaned$bleeding <- as.factor(Diagnosis_cleaned$bleeding)
Diagnosis_cleaned$fracture <- as.factor(Diagnosis_cleaned$fracture)
Diagnosis_cleaned$subdural_bleeding <- as.factor(Diagnosis_cleaned$subdural_bleeding)
Diagnosis_cleaned$epidural_bleeding <- as.factor(Diagnosis_cleaned$epidural_bleeding)
Diagnosis_cleaned$contusion <- as.factor(Diagnosis_cleaned$contusion)
Diagnosis_cleaned$other_bleeding <- as.factor(Diagnosis_cleaned$other_bleeding)
Diagnosis_cleaned$skull_base_fracture <- as.factor(Diagnosis_cleaned$skull_base_fracture)
Diagnosis_cleaned$skull_cap_fracture <- as.factor(Diagnosis_cleaned$skull_cap_fracture)
Diagnosis_cleaned$concussion <- as.factor(Diagnosis_cleaned$concussion)
Diagnosis_cleaned$brain_edema <- as.factor(Diagnosis_cleaned$brain_edema)
Diagnosis_cleaned$brain_compression <- as.factor(Diagnosis_cleaned$brain_compression)
Diagnosis_cleaned$concussion_correction_formula <- as.factor(Diagnosis_cleaned$concussion_correction_formula)
Diagnosis_cleaned$concussion_correction_including_unconsciousness_in_tbi <- as.factor(Diagnosis_cleaned$concussion_correction_including_unconsciousness_in_tbi)
Diagnosis_cleaned$unconsciousness <- as.factor(Diagnosis_cleaned$unconsciousness)
Diagnosis_cleaned$subarachnoid_bleeding <- as.factor(Diagnosis_cleaned$subarachnoid_bleeding)
Diagnosis_cleaned$diffuse_brain_injuries_and_ripping_shearing <- as.factor(Diagnosis_cleaned$diffuse_brain_injuries_and_ripping_shearing)


#save the df
save(Diagnosis_cleaned,
     file = paste0(clean_path, "Diagnosis_cleaned.RData"))
export(x = Diagnosis_cleaned,
       file = paste0(clean_path, "Diagnosis_cleaned.csv"))

```

# Codebook

* **research_case_id**: character, ID of the case (unique identifier).
* **date**: Date, date when the case or incident occurred (format: YYYY-MM-DD).
* **bleeding**: factor (binary), indicator for any bleeding present (0 = No, 1 = Yes).
* **fracture**: factor (binary), indicator for any fracture present (0 = No, 1 = Yes).
* **subdural_bleeding**: factor (binary), indicator for subdural bleeding (0 = No, 1 = Yes).
* **epidural_bleeding**: factor (binary), indicator for epidural bleeding (0 = No, 1 = Yes).
* **subarachnoid_bleeding**: factor (binary), indicator for subarachnoid bleeding (0 = No, 1 = Yes).
* **contusion**: factor (binary), indicator for contusion (0 = No, 1 = Yes).
* **diffuse_brain_injuries_and_ripping_shearing**: factor (binary), indicator for diffuse brain injuries and ripping/shearing (0 = No, 1 = Yes).
* **other_bleeding**: factor (binary), indicator for other types of bleeding not listed separately (0 = No, 1 = Yes).
* **skull_base_fracture**: factor (binary), indicator for skull base fracture (0 = No, 1 = Yes).
* **skull_cap_fracture**: factor (binary), indicator for skull cap (calvarium) fracture (0 = No, 1 = Yes).
* **unconsciousness**: factor (binary), indicator for unconsciousness present (0 = No, 1 = Yes).
* **concussion**: factor (binary), indicator for concussion (0 = No, 1 = Yes).
* **brain_edema**: factor (binary), indicator for brain edema (0 = No, 1 = Yes).
* **brain_compression**: factor (binary), indicator for brain compression (0 = No, 1 = Yes).
* **concussion_correction_formula**: factor (binary), derived variable for concussion correction based on internal logic (0 = No, 1 = Yes).
* **concussion_correction_including_unconsciousness_in_tbi**: factor (binary), derived variable for concussion correction including unconsciousness in TBI (0 = No, 1 = Yes).
* **injury_descriptions**: character, detailed narrative of all diagnosed injuries and conditions associated with the case, in German.
* **num_injuries**: integer, number of distinct injuries/diagnoses mentioned in *injury_descriptions*.





```{r Fast ispection}
# Histogram of the number of injuries per person
ggplot(Diagnosis_cleaned, aes(x = num_injuries)) +
  geom_bar(fill = 'lightgreen', color = 'black') +
  labs(title = 'Distribution of the Number of Injuries per Person', x = 'Number of Injuries', y = 'Count')



#Let's see how much each injury occurs
injury_columns <- c("bleeding", "fracture", "subdural_bleeding", "epidural_bleeding", 
                    "subarachnoid_bleeding", "contusion", "diffuse_brain_injuries_and_ripping_shearing",
                    "other_bleeding", "skull_base_fracture", "skull_cap_fracture", 
                    "unconsciousness", "concussion", "brain_edema", "brain_compression",
                    "concussion_correction_formula", "concussion_correction_including_unconsciousness_in_tbi")



#change to long format ( I dont know if necessary)
Diagnosis_long <- Diagnosis_cleaned %>%
  select(research_case_id, all_of(injury_columns)) %>%
  pivot_longer(cols = -research_case_id, names_to = "injury_type", values_to = "has_injury")


#Create table w percentages
summary_tab <- Diagnosis_long %>%
  group_by(injury_type) %>%
  summarise(percentage_injured = mean(has_injury == 1) * 100, .groups = 'drop')

# View the summarized table
summary_tab


##save the long fromat
save(Diagnosis_long,
     file = paste0(clean_path, "Diagnosis_long.RData"))
export(x = Diagnosis_long,
       file = paste0(clean_path, "Diagnosis_long.csv"))
```
# Codebook

* **research_case_id**: character, ID of the case (unique identifier), repeated for each injury type related to that case.
* **injury_type**: character, name of the specific injury or condition assessed for the case (e.g., "bleeding", "fracture", "subdural_bleeding").
* **has_injury**: integer (binary), indicator whether the specified injury/condition (*injury_type*) is present in the case (0 = No, 1 = Yes).




## Mapping Diagnosen

```{r mapping}
mapping <- import_dataset(paste0(import_path, 
                                "[SDS].[v_Export_Mapping_Diagnosen_ICD_to_AIS_tra_sds].csv"))

#First, drop columns "Displayname", "Name" (which are German ddescription of the injury), Kat research_id
mapping <- mapping %>%
  select(-Displayname, -Name, -Kat, -research_id)


sum(mapping$ICD_AIS_Code == mapping$Code)
nrow(mapping)

#ICD_AIS_Code and Code are exactly equal.
#As the other file, I'd drop Code (and so ICD_AIS_Code) and Version
mapping <- mapping %>%
  select(-ICD_AIS_Code, -Code, -Version)

# Convert "Date" column to Date type
mapping$Dat <- as.Date(mapping$Dat, format = "%Y-%m-%d %H:%M:%S")
```

```{r}
unique(mapping$AIS_CHAPTER) #note there is a -1
unique(mapping$ISS_BODY_REGION)
unique(mapping$MAX_AIS_SEVERITY)


#Maybe I am just tired but what are those? Anyway I'd convert to factor
"mapping$AIS_CHAPTER <- as.factor(mapping$AIS_CHAPTER) 
mapping$ISS_BODY_REGION <- as.factor(mapping$ISS_BODY_REGION)
mapping$MAX_AIS_SEVERITY <- as.factor(mapping$MAX_AIS_SEVERITY)"

#Missing?
table(is.na(mapping))



# Count NAs per column
na_count_per_column <- colSums(is.na(mapping))
na_count_per_column[na_count_per_column > 0]  # Display columns with NAs


# Calculate percentage of NA values per column
na_percentage <- colSums(is.na(mapping)) / nrow(mapping) * 100
na_percentage[na_percentage > 0]  # Display columns with NAs and their percentage

#the 0.3 % of the dates are missing.... so few. As before I will remove but we can always keep

na_any_column_df <- mapping[!complete.cases(mapping), ]


# Remove rows with NAs
mapping <- mapping[complete.cases(mapping), ]


# remove duplicate row
mapping <- unique(mapping)

```

```{r}
#Clean up the text data
mapping$DESCRIPTION <- tolower(mapping$DESCRIPTION)  # Convert to lowercase
mapping$DESCRIPTION <- gsub("\\s+", " ", mapping$DESCRIPTION)  # Remove extra spaces
mapping$DESCRIPTION <- trimws(mapping$DESCRIPTION)  # Remove leading and trailing space

# Frequency plot for DESCRIPTION
library(ggplot2)
top_display_names <- as.data.frame(table(mapping$DESCRIPTION))
colnames(top_display_names) <- c("display_name", "count")
top_display_names <- top_display_names[order(-top_display_names$count), ]
head(top_display_names, 10)  # Show top 10

ggplot(top_display_names[1:10, ], aes(x = reorder(display_name, -count), y = count)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(title = "Top 10 Most Common Display Names", x = "Display Name", y = "Count")

mapping_cleaned <- mapping


#save the df
save(mapping_cleaned,
     file = paste0(clean_path, "mapping_cleaned.RData"))
export(x = mapping_cleaned,
       file = paste0(clean_path, "mapping_cleaned.csv"))
```


# Codebook

* **AIS_CHAPTER**: integer, represents the chapter number of the AIS (Abbreviated Injury Scale) classification.
* **Dat**: Date, the date when the injury or event occurred (format: YYYY-MM-DD).
* **DESCRIPTION**: character, a detailed description of the injury or condition.
* **ISS_BODY_REGION**: integer, represents the body region affected by the injury based on AIS classification (for example, 4 could indicate a specific region like "limbs").
* **MAX_AIS_SEVERITY**: integer, the maximum severity score of the injury on the AIS scale (range 0 to 6).
* **research_case_id**: character, ID of the case, which uniquely identifies the patient or incident.


```{r}
# Check for duplicate patient IDs
dupes <- mapping %>%
  group_by(research_case_id) %>%
  filter(n() > 1)
print(paste("Number of duplicate patient entries:", nrow(dupes)))
```
So almost each patient has more than 1 row.
At this point there are so many ways to format the table to have a row for each patients that maybe we need to discuss about it. Here are some ideas: 

```{r}

# One row per patient
# Create a summary table with unique research_case_id for each row
patient_summary <- mapping %>%
  group_by(research_case_id) %>%
  summarise(
    
    # Get the first injury date for the patient
    latest_injury_date = min(Dat, na.rm = TRUE),
    
    # Calculate the number of distinct body regions injured
    distinct_body_regions_injured = n_distinct(ISS_BODY_REGION),
    
    # take  max severity per patient
    max_severity = max(MAX_AIS_SEVERITY, na.rm = TRUE),
    
    #OTHER IDEAS?? 
    
    # Sum of injuries 
    sum_injuries = sum(ISS_BODY_REGION != 0, na.rm = TRUE),  # don't count the zero
    
    # Concatenate all injury descriptions for the same patient
    all_injury_descriptions = paste(unique(DESCRIPTION), collapse = "; ")
  ) %>%
  ungroup()

# View the resulting table
head(patient_summary)



```


Or even something like that with the dummyes but idk
```{r}
patient_injury_counts_by_severity <- mapping %>%
  group_by(research_case_id) %>%
  summarise(
    count_severity_0 = sum(MAX_AIS_SEVERITY == 0, na.rm = TRUE),
    count_severity_1 = sum(MAX_AIS_SEVERITY == 1, na.rm = TRUE),
    count_severity_2 = sum(MAX_AIS_SEVERITY == 2, na.rm = TRUE),
    count_severity_3 = sum(MAX_AIS_SEVERITY == 3, na.rm = TRUE),
    count_severity_4 = sum(MAX_AIS_SEVERITY == 4, na.rm = TRUE),
    count_severity_5 = sum(MAX_AIS_SEVERITY == 5, na.rm = TRUE),
    count_severity_6 = sum(MAX_AIS_SEVERITY == 6, na.rm = TRUE),
  ) %>%
  ungroup()

```





