---
title: "Individual Models: Circulation, Temperature, INR vs. TBI"
output: html_document
date: "2025-04-15"
author: vjohner
---

```{r setup, include=FALSE}

rm(list = ls())

clean_path <- "02_data/02_clean_data/"
# for models?
# model_path <- "05_models/"

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(data.table)
library(lme4)      # For mixed effects models
library(lmerTest)  # For p-values in mixed models
library(ggplot2)   # For plotting
library(tidyverse)
library(car)       # For diagnostic plots
library(pROC)      # For ROC curves

knitr::opts_knit$set(root.dir = normalizePath(".."))
```

## Load and Prepare Data

```{r load_data}

poly <- readRDS(paste0(clean_path, "population_poly.rds"))

poly <- poly %>% select(research_case_id, bp, hr, temperature, inr, iss, iss_cat,
                        invasive, gcs, gcs_cat, sex, age, age_cat, bleeding, fracture,
                        concussion, brain_edema, brain_compression, unconsciousness)

# What to do with NA's?
# For heart rate and blood pressure must be documentation mistakes (always measured), so I exclude them in the analysis

poly <- poly %>% drop_na(bp, hr)

# Also, a HR of 865 seems hardly normal, take it out
poly <- poly %>% filter(hr < 250)

# For temp and inr missing probably indicates no problem and thus not measured. I will try to fit a model where I remove the NA's and another model where I convert the variables to ordered factors and classify NA's as in normal range (ASSUMPTION!).


# Create long-format dataset for BP and HR (hemorrhage)
poly_hemo <- melt(poly, 
                  id.vars = c("research_case_id", "gcs_cat", "iss_cat", "invasive", 
                              "sex", "age_cat", "bleeding", "fracture", "concussion", 
                              "brain_edema", "brain_compression", "unconsciousness"),
                  measure.vars = c("bp", "hr"),
                  variable.name = "vital_sign",
                  value.name = "value")

# Create combined GCS and vital sign variable
poly_hemo[, gcs_vital := paste(gcs_cat, vital_sign, sep = "_")]


```


## Exploratory Data Analysis

```{r explore_data}
# Distribution of ordered factors
table(poly$gcs_cat)
table(poly$iss_cat)
table(poly$age_cat)

# Summary statistics of key variables
summary(poly[, .(bp, hr, temperature, inr)])

# Check missing values in key variables
missing_values <- colSums(is.na(poly))
print(missing_values)

# Distribution of key outcome variables
par(mfrow=c(2,2))
hist(poly$bp, main="Blood Pressure Distribution", xlab="BP")
hist(poly$hr, main="Heart Rate Distribution", xlab="HR")
hist(poly$temperature, main="Temperature Distribution", xlab="Temperature")
hist(poly$inr, main="INR Distribution", xlab="INR")
par(mfrow=c(1,1))

# Examine relationships between ordered factors and outcomes
boxplot(bp ~ gcs_cat, data = poly, 
        main = "Blood Pressure by GCS Category", xlab = "GCS Category", ylab = "BP")
boxplot(hr ~ gcs_cat, data = poly,
        main = "Heart Rate by GCS Category", xlab = "GCS Category", ylab = "HR")
boxplot(temperature ~ gcs_cat, data = poly,
        main = "Temperature by GCS Category", xlab = "GCS Category", ylab = "Temperature")
boxplot(inr ~ gcs_cat, data = poly,
        main = "INR by GCS Category", xlab = "GCS Category", ylab = "INR")
```


## Model 1: Blood Pressure and Heart Rate (Hemorrhage)

```{r model_1_hemo}

model1 <- lmer(value ~ vital_sign * gcs_cat + iss_cat + invasive + sex + age_cat + 
               bleeding + fracture + concussion + brain_edema + brain_compression +
               unconsciousness + (1|research_case_id), data = poly_hemo)

# Alternative model with sex interaction
model1_sex <- lmer(value ~ vital_sign * gcs_cat + vital_sign * sex + iss_cat + invasive +
                   age_cat + bleeding + fracture + concussion + brain_edema +
                   brain_compression + unconsciousness + (1|research_case_id), 
                   data = poly_hemo)

# Model summaries
summary(model1)
summary(model1_sex)

# Compare models
anova(model1, model1_sex)

```

Note that following our discussion last time I also fit the model with an interaction with sex. However, neither the interaction the main effect were significant and it resulted in a higher AIC score. In the model without interaction, the main effect is significant, indicating higher bp and hr for men.

* vital_signhr (sign. negative): Heart rate is lower than blood pressure, all else equal
* vital_signhr:gcs_cat.L (sign. positive): This is the linear & quadratic interaction trend
between GCS and vital_sign = HR. A positive effect means that the effect of GCS category on HR increases linearly across levels.
* gcs_cat.L (sign. negative): Linear contrast for the GCS category. As severity increases (higher GCS category), the outcome (BP or HR) tends to decrease.
* gcs_cat.Q (sign. negative): Quadratic contrast. There’s a curve — e.g., the decrease might slow down or flatten at extremes.
* Unconsciousness (sign. negative): Negative effect on bp/hr

We can see a significant positive linear effect of age_cat.


Let's look at some diagnostics plots
```{r model_1_hemo_diag}

plot(model1) # Looks ok

qqnorm(resid(model1), main="Q-Q Plot of Residuals")
qqline(resid(model1))

# Half-normal plot of residuals (approximation using car package)
car::qqPlot(resid(model1), distribution="norm", envelope=0.95,
           main="Half-Normal Plot of Residuals", ylab="Residuals")

# Predictions
poly_hemo$predicted <- predict(model1, newdata = poly_hemo, allow.new.levels = TRUE, re.form = NULL)  # Include random effects

# Plot predicted vs observed
ggplot(poly_hemo, aes(x=value, y=predicted)) +
  geom_point(alpha=0.5) +
  geom_abline(intercept=0, slope=1, color="red") +
  facet_wrap(~vital_sign, scales="fixed") +
  labs(title="Predicted vs Observed Values", x="Observed", y="Predicted") +
  theme_minimal()

# Examine interaction effects with ordered factors
ggplot(poly_hemo, aes(x=gcs_cat, y=value, fill=vital_sign)) +
  geom_boxplot() +
  facet_wrap(~vital_sign, scales="free_y") +
  labs(title="BP and HR by GCS Category", x="GCS Category", y="Value") +
  theme_minimal()

```

The qq plot looks like there is a bit of skewness, we could try to fit it with a log trafo.

```{r model_1_hemo_log}
model1_log <- lmer(log(value + 0.0001) ~ vital_sign * gcs_cat + iss_cat + invasive + sex +
               age_cat + bleeding + fracture + concussion + brain_edema +
               brain_compression + unconsciousness + (1|research_case_id), 
               data = poly_hemo)

summary(model1_log)

anova(model1, model1_log)

# Some Diagnostics
plot(model1_log)

qqnorm(resid(model1_log), main="Q-Q Plot of Residuals")
qqline(resid(model1_log))

# Half-normal plot of residuals (approximation using car package)
car::qqPlot(resid(model1), distribution="norm", envelope=0.95,
           main="Half-Normal Plot of Residuals", ylab="Residuals")

# Predictions
poly_hemo$predicted_log <- predict(model1_log, newdata = poly_hemo, allow.new.levels = TRUE, re.form = NULL)  # Include random effects

poly_hemo$predicted_log_trafo <- exp(poly_hemo$predicted_log) - 0.0001

# Plot predicted vs observed
ggplot(poly_hemo, aes(x=value, y=predicted_log_trafo)) +
  geom_point(alpha=0.5) +
  geom_abline(intercept=0, slope=1, color="red") +
  facet_wrap(~vital_sign, scales="fixed") +
  labs(title="Predicted vs Observed Values", x="Observed", y="Predicted") +
  theme_minimal()

```
Changes in interpretation:
* Sex & unconsciousness is not significant anymore
* ISS not only quadratic barely significant but both linear and quadratic highly significant (negative), which is to be expected

At least to me, qqplot and predictions look better with a log trafo!

However, we can see a few outliers. What worries me is that they did not show before the log trafo. We can have a look at them:
```{r model_1_outliers}
# TODO

# wrong ones before

```

* What should I do about these? Some trafo? Robust regression? *
* Note that maybe log trafo not so good, before these "outliers" did not show*



## Model 2: Temperature as Outcome

Note that since we only have temperature (single outcome variable; continuous) and no multivariate case here, there's no point in fitting a lmer() model, since it's not sensible to have random effects for research_case_id (here, they are unique!).

```{r model_2_temp}
poly_temp <- poly %>% drop_na(temperature)

qqnorm(poly_temp$temperature, main="Q-Q Plot for Temperature")
qqline(poly_temp$temperature)

model2 <- lm(temperature ~ gcs_cat + iss_cat + invasive + sex + age_cat + 
             bleeding + fracture + concussion + brain_edema + brain_compression +
             unconsciousness, data = poly_temp)

summary(model2)
# Figure out what to do with NA --> make one model with removed
# --> one model with categories

```
 Interpretation:
 
 * Higher GCS, higher ISS, and older age predict lower log temperature.
 * Bleeding and concussion are associated with higher log temperature.
 * Male not significant.

Let's look at some diagnostics:
```{r model_2_temp_diag}

par(mfrow=c(2,2))
plot(model2)
par(mfrow=c(1,1))

# Half-normal plot of residuals
car::qqPlot(resid(model2), distribution="norm", envelope=0.95,
           main="Half-Normal Plot of Residuals", ylab="Residuals")

# Predictions
poly_temp$temp_predicted <- predict(model2, poly_temp)

# Plot predicted vs observed
ggplot(poly_temp, aes(x=temperature, y=temp_predicted)) +
  geom_point(alpha=0.5) +
  geom_abline(intercept=0, slope=1, color="red") +
  labs(title="Predicted vs Observed Temperature", x="Observed", y="Predicted") +
  theme_minimal()

```

I don't know, doesn't look so good.

**Try with ordered factors instead of continuous temperature! (see below)**


Just look at effects:
``` {r effects_temp}
# Examine effects of ordered factors
ggplot(poly_temp, aes(x=gcs_cat, y=temperature)) +
  geom_boxplot() +
  labs(title="Temperature by GCS Category", x="GCS Category", y="Temperature") +
  theme_minimal()

ggplot(poly_temp, aes(x=iss_cat, y=temperature)) +
  geom_boxplot() +
  labs(title="Temperature by ISS Category", x="ISS Category", y="Temperature") +
  theme_minimal()

ggplot(poly_temp, aes(x=age_cat, y=temperature)) +
  geom_boxplot() +
  labs(title="Temperature by Age Category", x="Age Category", y="Temperature") +
  theme_minimal()

```


## Model 2.2: Temperature as Outcome (but w/ ordered factors)















## Model 3: INR as Outcome

Same thing as above; just use lm.

```{r model_3_inr}
poly_inr <- poly %>% drop_na(inr)

model3 <- lm(inr ~ gcs_cat + iss_cat + invasive + sex + age_cat + 
             bleeding + fracture + concussion + brain_edema + brain_compression +
             unconsciousness, data = poly_inr)

summary(model3)

```

Let's look at some diagnostics:
```{r model_3_diagnostics}
par(mfrow=c(2,2))
plot(model3)
par(mfrow=c(1,1))

# Examine relationship between INR and ordered factors
ggplot(poly, aes(x=gcs_cat, y=inr)) +
  geom_boxplot() +
  labs(title="INR by GCS Category", x="GCS Category", y="INR") +
  theme_minimal()

summary(poly$inr)
boxplot(poly$inr)
boxplot(log(poly$inr))
```

Very bad. If we look at the last plots it's no wonder, a lot of outliers or very high values distorting it. **Maybe approach with ordered factors can help.**



Try with a log trafo:
```{r model_3_inr_log}

model3_log <- lm(log(inr + 0.0001) ~ gcs_cat + iss_cat + invasive + sex + age_cat + 
                 bleeding + fracture + concussion + brain_edema + brain_compression +
                 unconsciousness, data = poly_inr)

summary(model3_log)

par(mfrow=c(2,2))
plot(model3_log)
par(mfrow=c(1,1))

```

Same result. Helps a bit, but not nearly enough.












