---
title: "Model: Four Dimensions of polytrauma pathophysiology vs. TBI"
output: html_document
date: "2025-04-01"
author: jaegerl
---

```{r setup, include=FALSE}
# define paths
# for data import
import_path <- "02_data/01_raw_data/"
# for export of inspection objects (spreadsheets)
inspect_path <- "03_spreadsheets/"
# for export of clean data
clean_path <- "02_data/02_clean_data/"
# for intermediate results
intermediate_path <- "02_data/03_dump/"
# for tables
table_path <- "04_tables/"
# for figures
figure_path <- "05_figures/"

knitr::opts_chunk$set(echo = TRUE)

library(data.table)
library(rio)
library(tableone)
library(vcd)
library(poLCA)
library(glmmTMB)
library(tidyverse)

knitr::opts_knit$set(root.dir = normalizePath(".."))
```


## Build and Explore Dataset

Do we even see differences in pathophysiology by, say, degree of TBI?

Make a population dataset first.

```{r explore1}
# patient population and TBI diagnoses
load(paste0(clean_path, "Diagnosis_cleaned.RData"))
setDT(Diagnosis_cleaned)
# demographics, hospital stay, ICU stay
load(paste0(clean_path, "patients_clean.RData"))
# pathophysiology at admission
load(paste0(clean_path, "vitals_clean.RData"))
load(paste0(clean_path, "lab_clean.RData"))
# trauma scores
load(paste0(clean_path, "iss_clean.RData"))
load(paste0(clean_path, "mapping_cleaned.RData"))
# interventions
load(paste0(clean_path, "interventions1_wide.RData"))

# get demographics into the population
diagnoses <- c("bleeding", "fracture", "concussion",
               "brain_edema", "brain_compression",
               "unconsciousness")
Diagnosis_cleaned[, (diagnoses) := lapply(.SD, \(x) as.integer(as.character(x))),
                  .SDcols = diagnoses]
population <- merge(Diagnosis_cleaned[, .(research_case_id,
                                          date,
                                          bleeding,
                                          fracture,
                                          concussion,
                                          brain_edema,
                                          brain_compression,
                                          unconsciousness)],
                    y = patients_clean[, .(research_case_id,
                                           age, elderly, sex,
                                           stay_days, icu, icu_days,
                                           death, death_days)],
                    by = "research_case_id",
                    all.x = TRUE)
# identify those with TBI in a separate dataset with the diagnoses
tbi <- Diagnosis_cleaned[bleeding + fracture + concussion + 
                           brain_edema + brain_compression +
                           unconsciousness > 0]
population[, tbi := as.integer(research_case_id %in% tbi$research_case_id)]
# get vitals of patients
population <- merge(x = population,
                    y = vitals_clean,
                    by = "research_case_id",
                    all.x = TRUE)
# get lab data of patients
population <- merge(x = population,
                    y = lab_clean_wide,
                    by = "research_case_id",
                    all.x = TRUE)
# get ISS of patients
population <- merge(x = population,
                    y = iss_clean,
                    by = "research_case_id",
                    all.x = TRUE)

# calculate region AIS of each patient as maximum AIS value for that region
ais <- mapping_cleaned[, .(ais = max(MAX_AIS_SEVERITY)),
                       by = .(research_case_id,
                              AIS_CHAPTER)]
# define the names of the ais regions
ais_regions <- data.table(
  AIS_CHAPTER = -1:9,
  ais_region = c("No Map", "Other Trauma",
                 "Head", "Face", "Neck",
                 "Thorax", "Abdomen", "Spine",
                 "Upper Extremity", "Lower Extremity", "Unspecified"))
ais <- merge(ais,
             y = ais_regions,
             by = "AIS_CHAPTER",
             all.x = TRUE)
table(ais$ais_region)
# convert to a wide table
ais <- dcast(ais, research_case_id ~ ais_region, value.var = "ais")
ais <- ais[, mget(c("research_case_id", ais_regions$ais_region))]
# merge into population data
population <- merge(population,
                    ais,
                    by = "research_case_id",
                    all.x = TRUE)
# identify those with missing values
population[, `:=`(novital = as.integer(!(research_case_id %in% 
                                           vitals_clean$research_case_id)),
                  nolab = as.integer(!research_case_id %in%
                                       lab_clean_wide$research_case_id),
                  noais = as.integer(!(research_case_id %in%
                                         ais$research_case_id)),
                  noiss = as.integer(is.na(iss)))]

# identify patients with polytrauma as AIS > 0 in at least two body regions
polytrauma <- mapping_cleaned[MAX_AIS_SEVERITY > 0 & AIS_CHAPTER != 1
][, .(n_regions = n_distinct(AIS_CHAPTER)),
  by = .(research_case_id)]
population <- merge(population,
                    y = polytrauma,
                    by = "research_case_id",
                    all.x = TRUE)
population[is.na(n_regions), n_regions := 0]
population[, polytrauma := as.integer(n_regions >= 2)]

# define pathophysiology indicators
population[, `:=`(hypotension = as.integer(bp < 60),
                  tachycardia = as.integer(hr > 100),
                  acidosis = as.integer(lactate > 2),
                  severethoracic = as.integer(Thorax >= 2),
                  majortrauma = as.integer(iss >= 15),
                  anemia = as.integer(hemoglobin < 80),
                  coagulopathy = as.integer(inr >= 1.4),
                  hypothermia = as.integer(temperature < 35))]
# I guess that for all of these indicators, not measured can be regarded as 
# absent...
pph <- c("hypotension", "tachycardia", "anemia",
         "severethoracic", "majortrauma",
         "acidosis", "coagulopathy", 
         "hypothermia")
population[, (pph) := lapply(.SD, \(x) ifelse(is.na(x), 0, x)), .SDcols = pph]

# categorize GCS
population[, gcs_cat := case_when(
  is.na(gcs) ~ "0.unknown",
  gcs >= 13 ~ "1.mild",
  gcs >= 9 ~ "2.moderate",
  gcs <=8 ~ "3.severe")]

# identify presence of invasive intervention
interventions1_wide[, invasive := as.integer(
  rowSums(interventions1_wide[, .SD, .SDcols = !c(1, 2)]) > 0)]
population <- merge(x = population,
                    y = interventions1_wide[, .(research_case_id,
                                                invasive)],
                    by = "research_case_id",
                    all.x = TRUE)
# as discussed with the client, those without a record of interventions simply
# had no intervention
population[is.na(invasive), invasive := 0]

# define age categories
population[, age_cat := case_when(
  age < 40 ~ "1.<40",
  age < 60 ~ "2.40-59",
  age < 80 ~ "3.60-78",
  TRUE ~ "4.79+")]

# define pathophysiology variables
pph <- c("hypotension", "tachycardia", "anemia",
         "severethoracic", "majortrauma",
         "acidosis", "coagulopathy",
         "hypothermia")
# define injuries
injuries <- c("bleeding",
              "fracture",
              "concussion",
              "brain_edema",
              "brain_compression",
              "unconsciousness")
# define relevant patient characteristics
characteristics <- c("gcs_cat", 
                     "sex", 
                     "age_cat",
                     "invasive")

pph_data <- population[, c("research_case_id",
                           pph,
                           injuries,
                           characteristics), with = FALSE]
pph_data <- melt(pph_data, 
                 measure.vars = pph, variable.name = "pph",
                 value.name = "altered",
                 id.vars = c("research_case_id",
                             injuries,
                             characteristics)
                 )[order(research_case_id)]

# define the four dimensions of pathophysiology
pph_data[, pph_dim := case_when(
  pph %in% c("hypotension", "tachycardia", "anemia") ~ "shock",
  pph == "coagulopathy" ~ "coagulopathy",
  pph %in% c("severethoracic", "majortrauma", "acidosis") ~ "softtissue",
  pph == "hypothermia" ~ "hypothermia")]
dimensions <- c("shock", "coagulopathy", "softtissue", "hypothermia")

CreateTableOne(data = population,
               vars = c(pph, "gcs_cat"),
               strata = "tbi",
               factorVars = pph) |> 
  print(test = FALSE, noSpaces = TRUE)
```
It does look like there are several differences in the pathophysiology of polytrauma between patients with and without TBI:

* Signs of shock (hypothension, tachycardia) are more common in TBI.
* Coagulopathy is more common in TBI.
* Severe soft tissue injuries (as expressedy by total ISS and severe thoracic injury) are more common in TBI. Acidosis is much more common in TBI.
* Hypothermia is more common in TBI.

On the other hand, it is not clear if these differences are simply an expression of TBI cases being involved in more severe polytrauma, as may be indicated by the higher ISS in TBI cases.

Also note that the single alterations in pathophysiology measures can be very rare (esp. hypothermia). It may make sense to merge the categories for shock and soft tissue injury by logical dijunction, that is, `shock` as `hypotension | tachycardia | anemia` and `softtissue` as `severethoracic | majortrauma | acidosis`.

```{r explore2}
population[, `:=`(shock = as.integer(hypotension | tachycardia | anemia),
                  softtissue = as.integer(severethoracic | majortrauma | acidosis),
                  coagulopathy = as.integer(coagulopathy),
                  hypothermia = as.integer(hypothermia))]
CreateTableOne(data = population,
               vars = dimensions,
               strata = "tbi",
               factorVars = dimensions) |> 
  print(test = FALSE, noSpaces = TRUE)
```
Let us see if alterations in the different dimensions of pathophysiology differ by ISS. Note that soft tissue injury occurs by definition whenever ISS is at least 16. When adjusting for ISS, it may make sense to remove the ISS criterion from soft tissue injury.


```{r explore3}
# define ISS categories
population[, softtissue2 := as.integer(severethoracic | acidosis)]
population[, iss_cat := case_when(
  iss <= 8 ~ "1.1-8",
  iss <= 15 ~ "2.9-15",
  iss <= 24 ~ "3.16-24",
  iss >= 25 ~ "4.25+")]
table(population[polytrauma == 1, (iss_cat)], useNA = "always")
CreateTableOne(data = population[polytrauma == 1],
               vars = gsub(x = dimensions, 
                           pattern = "softtissue",
                           replacement = "softtissue2"),
               strata = "iss_cat",
               factorVars = c(dimensions, "softtissue2")) |> 
  print(test = FALSE, noSpaces = TRUE)
```
The severer the polytrauma, the more common the alterations in the different dimensions of pathophysiology. 

## A First Model

Fit a first multivariate logistic regression model that relates alterations in the different dimensions of pathophysiology with the different patient characteristics, including GCS at baseline (categorized), age (categorized), sex, presence of invasive surgery, and the different types of injuries (bleeding, fracture, etc.). The outcome variable `altered` is binary and denotes whether the corresponding pathophyisiology parameter is altered. The model includes a random intercept for each case The model is fitted using the `glmmTMB` package.


```{r pph1}
dimensions <- c("shock", "coagulopathy", "softtissue2", "hypothermia")
characteristics <- c("gcs_cat", 
                     "iss_cat",
                     "sex", 
                     "age_cat",
                     "invasive")
pph_data <- population[, c("research_case_id",
                           dimensions,
                           injuries,
                           characteristics), with = FALSE]
pph_data <- melt(pph_data, 
                 measure.vars = dimensions, 
                 variable.name = "dimension",
                 value.name = "altered",
                 id.vars = c("research_case_id",
                             injuries,
                             characteristics)
                 )[order(research_case_id)]
pph_model1 <- glmmTMB(altered ~ gcs_cat + invasive + iss_cat + sex + age_cat +
                        (1 | research_case_id),
                      data = pph_data,
                      family = "binomial")
summary(pph_model1)
```

More complex version that may be more appropriate for the research question: Allow different coefficients for TBI by pathophysiology dimension:

```{r pph1_2}
pph_model1_2 <- glmmTMB(altered ~ gcs_cat*dimension + invasive + 
                          iss_cat + sex + age_cat +
                          (1 | research_case_id),
                        data = pph_data,
                        family = "binomial")
summary(pph_model1_2)
```

This looks promising!

Note that this first model involved the entire population of patients -- we will have to restrict to those with polytrauma only.

After supervisor meeting: 

* Use more categories and model ordinal variables as ordered factors.
* Perform some model diagnostics (half-normal plots, RE plots, etc.).
* Fit single models to the different outcomes.
