---
title: 'Ordinal Model: ISS, AIS thorax, AIS Head'
author: "Saverio Fontana"
date: "2025-04-15"
output: html_document
---

```{r setup, include=FALSE}
rm(list = ls())

clean_path <- "02_data/02_clean_data/"
# for models?
# model_path <- "05_models/"

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(MASS)
library(data.table)
library(lme4)      # For mixed effects models
library(lmerTest)  # For p-values in mixed models
library(ggplot2)   # For plotting
library(tidyverse)
library(car)       # For diagnostic plots
library(pROC)      # For ROC curves
library(ordinal)


knitr::opts_knit$set(root.dir = normalizePath(".."))
```



```{r load_data}
poly <- readRDS(paste0(clean_path, "population_poly.rds"))

# Distribution of ordered factors
table(poly$gcs_cat)
table(poly$iss_cat)
#very imbalanced
```



```{r}
# Check missing values in key variables
missing_summary <- sapply(poly, function(x) sum(is.na(x)))
missing_summary[missing_summary > 0]  # Show only variables with NAs
```

I am going to model iss_cat with a polr. No missing in it!


```{r}
library(ordinal)
model <- clm(iss_cat ~ gcs_cat + invasive + sex + age_cat + 
                 bleeding + fracture + concussion + brain_edema + brain_compression +
                 unconsciousness, data = poly)
summary(model)
```
Need to check PO assumption before considering the model reliable:


Visual inspection and AIC BIC is what I will use. 



First we specify two variables with non proportional odds. Can be for example.
```{r}

# Example:  non-proportional odds for 'bleeding' and 'fracture'
model_clm <- clm(iss_cat ~ gcs_cat + invasive + sex + age_cat +
                     concussion + brain_edema + brain_compression +
                     unconsciousness,
                   data = poly,
                   nominal = ~ bleeding + fracture) # Specify variables with non-proportional odds

summary(model_clm)
```



Compare model to see if sign better the fit

```{r}
anova(model, model_clm)

#Probably violated then, so we need the second 
```


Visual inspection:

```{r}
library(broom)
library(ggplot2)

# Extract coefficients and filter out threshold terms
coef_data <- tidy(model_clm) %>%
  filter(!grepl("\\|", term)) %>%  # Remove threshold terms (e.g., "1|2", "2|3")
  mutate(
    term = factor(term, levels = term[order(estimate)]),  # Order terms by estimate
    significance = ifelse(p.value < 0.05, "Significant", "Non-significant")
  )

# Plot
ggplot(coef_data, aes(x = estimate, y = reorder(term, estimate), color = significance)) +
  geom_point(size = 3) +
  geom_errorbarh(
    aes(xmin = estimate - 1.96 * std.error, xmax = estimate + 1.96 * std.error),
    height = 0.2
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  scale_color_manual(values = c("Significant" = "blue", "Non-significant" = "gray")) +
  labs(
    title = "Coefficient Estimates with 95% Confidence Intervals",
    x = "Log-Odds Estimate",
    y = "Predictor",
    color = "Significance"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```


```{r}
tidy(model_clm) %>%
  filter(grepl("bleeding|fracture", term)) %>%  
  separate(term, into = c("variable", "threshold"), sep = "\\|") %>%  
  mutate(
    variable = factor(variable),  
    threshold = factor(threshold, levels = unique(threshold)),  # Keep original threshold order
    # Create a combined factor with custom ordering
    term_group = fct_rev(interaction(variable, threshold, sep = " | ", drop = TRUE))  
  ) %>%  
  ggplot(aes(
    x = estimate,
    y = term_group,  
    color = variable  # Color by variable for visual grouping
  )) +
  geom_point() +
  geom_errorbarh(aes(
    xmin = estimate - 1.96 * std.error,
    xmax = estimate + 1.96 * std.error
  )) +
  labs(
    color = "Variable",
    y = "Term (Variable | Threshold)",
    x = "Estimate with 95% CI"
  ) +
  scale_y_discrete(labels = function(x) gsub("^.*\\|", "", x)) +  # Show only threshold on y-axis
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 9),
    legend.position = "bottom"
  )
```

 The model estimates cumulative log odds of being in a certain category or below vs. being in higher categories
 
 
 Positive coefficient → increases odds of being in a lower severity category.

Negative coefficient → increases odds of being in a higher severity category



gcs_cat: Higher GCS → Higher odds of being in a more severe ISS category.
Quadratic (Q) and Cubic (C) terms also significant, suggesting non-linear effects.

Sex: Males have higher odds of severe ISS than females.

Age::
LINEAR: n general, increasing age is associated with lower injury severity (ISS). -> Older patients are less likely to be in the highest ISS categories compared to younger patients.

Possible reasons:
Survivorship bias (severe trauma is more lethal in older adults, so survivors have lower ISS).
Older adults may experience less high-energy trauma (e.g., fewer car crashes, more low-energy falls).

QUADRATIC: U-SHAPED relationship
Middle-aged groups (e.g., 30–50 years) have the lowest ISS severity (bottom of the U-curve).
Both younger and older patients trend toward higher severity

CUBIC: not sign

QUARTIC 
Interpretation: A positive coefficient adds a "W-shape" to the U-curve, suggesting:
Subgroups within age categories may have distinct risk patterns.
Example: A "bump" in severity for very young (18–25) and very old (80+) patients.

other terms not sign


Concussion: Concussion → Lower odds of severe ISS (counterintuitive; may need clinical context).

Invasive Procedures, Brain Edema, Brain Compression, and Unconsciousness: These factors did not show statistically significant associations with injury severity in this model.


NON PROP:
INTERCEPT: boundaries between ISS categories *when all predictors = 0* (reference groups):

1.1-8|2.9-15: 0.063 (not significant, p ≈ 0.37)
Weak separation between mild (1.1-8) and moderate (2.9-15) ISS.

2.9-15|3.16-24: 1.773 (p < 0.001)
Strong separation between moderate and severe ISS.

3.16-24|4.25+: 3.341 (p < 0.001)
Very strong separation between severe and critical ISS.




Bleeding: Across all thresholds, bleeding increases the probability of being in more severe categories, with the largest effect at the lowest threshold (presence of bleeding strongly lowers the threshold between categories 1 and 2)

Fracture: 
1.1-8|2.9-15: A very strong effect: having a fracture greatly increases the odds of being in category 2 or higher, i.e., shifts you away from the least severe category.
2.9-15 vs 3.16-24	-0.74	Also increases odds of being in category 3 or higher (vs. category 2), though the effect is smaller.
3.16-24 vs 4.25+	+0.28	Opposite direction! This raises the threshold, making it slightly less likely to be in the most severe category (category 4), all else equal.




I tried other visual inspecion:

```{r}
model_clm_extended <- clm(iss_cat ~ invasive + sex + 
                                 concussion + brain_edema +
                                brain_compression,
                              data = poly,
                              nominal = ~ bleeding + fracture + gcs_cat +
                                        invasive + age_cat + unconsciousness)

summary(model_clm_extended)

anova(model_clm, model_clm_extended) #slightly better
```
Sex: Male patients have a significantly higher likelihood of being in a more severe injury category compared to female patients, and this effect is consistent across all severity levels.
Concussion: The presence of a concussion is associated with a significantly lower likelihood of being in a more severe injury category, and this effect is consistent across all severity levels.


Bleeding: The presence of bleeding significantly increases the likelihood of being in a higher injury severity category compared to being in a lower one at all severity thresholds. The strongest effect is in distinguishing the lowest severity injuries from more severe ones.
Fracture: Fractures significantly increase the likelihood of being in higher severity categories compared to the lowest. Additionally, having a fracture significantly increases the likelihood of being in the highest severity category compared to the middle ones.
GCS Category: More severe neurological impairment (as indicated by gcs_cat) is significantly associated with a higher likelihood of being in more severe injury categories at all thresholds. Specifically, the linear trend (gcs_cat.L) shows a significant negative association with being above each severity threshold, indicating that worsening GCS leads to higher injury severity. There are also significant non-linear effects, with the cubic trend (gcs_cat.C) showing a significant positive association with being above the middle and highest severity thresholds, and the quadratic trend (gcs_cat.Q) showing a significant negative association at the lowest threshold.
Invasive Procedures: Undergoing an invasive procedure significantly decreases the likelihood of being in the lowest injury severity category compared to higher categories.
Age Category:
The linear trend of age (age_cat.L) shows a significant positive association with being above the middle and highest severity thresholds, suggesting that older age is linked to higher injury severity at these levels.
The quadratic trend of age (age_cat.Q) also shows a significant positive association with being above the lowest and middle severity thresholds.
The 4th-order polynomial of age (age_cat^4) shows a significant negative association with being above the lowest severity threshold.
Unconsciousness: Being unconscious significantly decreases the likelihood of being in the lowest injury severity category compared to higher categories.




Well the model actually improves...



```{r}
tidy(model_clm_extended) %>%
  filter(grepl("bleeding|fracture|gcs_cat|invasive|age_cat|unconsciousness", term)) %>%  
  separate(term, into = c("variable", "threshold"), sep = "\\|") %>%  
  mutate(
    variable = factor(variable),  
    threshold = factor(threshold, levels = unique(threshold)),  # Keep original threshold order
    # Create a combined factor with custom ordering
    term_group = fct_rev(interaction(variable, threshold, sep = " | ", drop = TRUE))  
  ) %>%  
  ggplot(aes(
    x = estimate,
    y = term_group,  
    color = variable  # Color by variable for visual grouping
  )) +
  geom_point() +
  geom_errorbarh(aes(
    xmin = estimate - 1.96 * std.error,
    xmax = estimate + 1.96 * std.error
  )) +
  labs(
    color = "Variable",
    y = "Term (Variable | Threshold)",
    x = "Estimate with 95% CI"
  ) +
  scale_y_discrete(labels = function(x) gsub("^.*\\|", "", x)) +  # Show only threshold on y-axis
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 9),
    legend.position = "bottom"
  )
```
Even though the extended model is significantly bettrer, I decided to keep the easier model only with bleedingand fracture as treshold varying cause the other variables, by visual inspection, have a varying treshold coefficient which is very similar (even if sometimes they are sign different at the 95) with overlapping CI and especially, the sign is the same, while for the first two, the sign is different.



## Final fit





```{r}
summary(model_clm)
```




























These are some extensions tot he analysis
```{r}
nominal_test(model)  # Function directly from the ordinal package
```



```{r}
#evaluate prediction, but not relevant
library(caret)
predicted <- predict(model_clm, type = "class")$fit
caret::confusionMatrix(predicted, poly$iss_cat)
```



##############################
#ISS Continuotus
```{r}
hist(poly$iss)
summary(poly$iss) #a bit skewed
```

```{r}
model_lm <- lm(iss ~ sex + 
                                 concussion + brain_edema +
                                brain_compression + bleeding + fracture + gcs_cat +
                                        invasive + age_cat + unconsciousness,
                              data = poly)
summary(model_lm)
```



```{r}
plot(model_lm)
```


 to have a slightly better QQ-plot:
```{r}
model_log <- lm(log(iss) ~ sex + 
                                 concussion + brain_edema +
                                brain_compression + bleeding + fracture + gcs_cat +
                                        invasive + age_cat + unconsciousness,
                              data = poly)
summary(model_log)
plot(model_log)

```
All effects are interpreted multiplicatively on the ISS scale by exponentiating the coefficients.

Sex (Male): Being male is associated with a 9.7% higher ISS on average compared to females, holding other variables constant (exp(0.093) ≈ 1.097).

Concussion: The presence of a concussion is associated with a 14% decrease in ISS (exp(−0.150) ≈ 0.861), suggesting a protective or inverse relationship with severity.

Bleeding: Bleeding is associated with a 59.5% increase in ISS on average (exp(0.467) ≈ 1.595), indicating it is a strong predictor of higher injury severity.

Fracture: The presence of a fracture increases ISS by approximately 53.4% (exp(0.428) ≈ 1.534), also marking it as a strong positive predictor.

GCS Score (gcs_cat):

The linear trend (gcs_cat.L) shows that worse GCS scores (lower neurological status) are associated with 35.6% higher ISS (exp(0.305) ≈ 1.356).

The other orthogonal polynomial terms (.Q, .C) suggest some non-linear effects but with smaller or less consistent influence.

Age (age_cat):

Only the second-order polynomial term (age_cat.Q) is statistically significant, showing that middle age groups may have 14.4% lower ISS (exp(−0.155) ≈ 0.857) relative to the reference trend, but the effect is non-linear and should be interpreted cautiously (IT WAS A BIT DIFFERENT FROM THE ORDINAL).

Unconsciousness: Being unconscious is associated with a 22.8% higher ISS (exp(0.205) ≈ 1.228), reflecting greater injury severity. (IN THE ORDINAL IT WASNT SIGNIFICANT)

Invasive procedures, brain edema, and brain compression did not show statistically significant associations with ISS in this model.




# What if box and cox trafo??
```{r}
boxcox(model_lm, lambda = seq(-2, 2, 0.1))
bc <- boxcox(model_lm, lambda = seq(-2, 2, 0.1), plotit = FALSE)
optimal_lambda <- bc$x[which.max(bc$y)]
optimal_lambda
```

 Use lambda = 0.3, since it's close to zero, instead of doing y^0.3, I would stick with log for interpretability. 


# Residuals vs predictors
```{r}
for (var in c("gcs_cat", "invasive", "sex", 
              "bleeding", "fracture", "concussion", "brain_edema", 
              "brain_compression", "unconsciousness")) {
  plot(poly[[var]], resid(model_lm), main = paste("Residuals vs", var), xlab = var, ylab = "Residuals")
}

```


















Robust regression makes any sense? 

```{r}
library(robustbase)
model_rob <- lmrob(iss ~ sex + 
                                 concussion + brain_edema +
                                brain_compression + bleeding + fracture + gcs_cat +
                                        invasive + age_cat + unconsciousness,
                              data = poly, fast.s.large.n = Inf)
summary(model_rob)
plot(model_rob)
```

log + robust?? 



Or adding the variable --> severe_thoracic_injury. I dont think it makes sense: iss is made up of AIS, so of course they are highly correlated


################################
#TRY WITH CK (continuous outocome)
Now I dont want to change anything here, I guess the results are similar to Vale's, so just wait for his questions and then implement what he found. Cause there are many missings and many outliers!
Robreg + imputation maybe



  
```{r}
#COUNT CK NAS
sum(is.na(poly$ck))
#DROP nas
poly_ck <- poly %>% drop_na(ck)

# Examine effects of ordered factors
ggplot(poly_ck, aes(x=gcs_cat, y=ck)) +
  geom_boxplot() +
  labs(title="Ck by GCS Category", x="GCS Category", y="Ck") +
  theme_minimal()

ggplot(poly_ck, aes(x=iss_cat, y=ck)) +
  geom_boxplot() +
  labs(title="Ck by ISS Category", x="ISS Category", y="CK") +
  theme_minimal()

ggplot(poly_ck, aes(x=age_cat, y=ck)) +
  geom_boxplot() +
  labs(title="Ck by Age Category", x="Age Category", y="CK") +
  theme_minimal()
```
 
it's highly skewed. Maybe we need better ispection. 
```{r}
qqnorm(poly_ck$ck, main="Q-Q Plot") #mh

#fit a lm anyway 
model <- lm(ck ~ gcs_cat + iss_cat + invasive + sex + age_cat + 
             bleeding + fracture + concussion + brain_edema + brain_compression +
             unconsciousness, data = poly_ck)

summary(model)

```

```{r}
par(mfrow=c(2,2))
plot(model)
par(mfrow=c(1,1))

# Half-normal plot of residuals
car::qqPlot(resid(model), distribution="norm", envelope=0.95,
           main="Half-Normal Plot of Residuals", ylab="Residuals")

# Predictions
poly_ck$ck_predicted <- predict(model, poly_ck)

# Plot predicted vs observed
ggplot(poly_ck, aes(x=ck, y=ck_predicted)) +
  geom_point(alpha=0.5) +
  geom_abline(intercept=0, slope=1, color="red") +
  labs(title="Predicted vs Observed CK", x="Observed", y="Predicted") +
  theme_minimal()
```

I have these outliers that really make the fit very difficult. Should I remove them? Clinically speaking are they possible

Inspection of the outliers

```{r}
summary(poly_ck$ck)

boxplot(poly_ck$ck, horizontal = TRUE, main = "Boxplot of CK Levels", col = "lightblue")

Q1 <- quantile(poly_ck$ck, 0.25, na.rm = TRUE)
Q3 <- quantile(poly_ck$ck, 0.75, na.rm = TRUE)
IQR <- Q3 - Q1
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

outliers <- poly_ck$ck[poly_ck$ck < lower_bound | poly_ck$ck > upper_bound]
print(paste("Number of outliers:", length(outliers)))
print(outliers)
```

Ok I have two ideas:
1) Censor the 99th percentile 
```{r}
upper_limit <- quantile(poly_ck$ck, 0.99, na.rm = TRUE)
poly_ck$ck_cut<- ifelse(poly_ck$ck > upper_limit, upper_limit, poly_ck$ck)
summary(poly_ck$ck_cut)
```

```{r}
qqnorm(poly_ck$ck_cut, main="Q-Q Plot") #mh

model2 <- lm(ck_cut ~ gcs_cat + iss_cat + invasive + sex + age_cat + 
             bleeding + fracture + concussion + brain_edema + brain_compression +
             unconsciousness, data = poly_ck)

summary(model2)

anova(model, model2)
```
```{r}
par(mfrow=c(2,2))
plot(model2)
par(mfrow=c(1,1))

# Half-normal plot of residuals
car::qqPlot(resid(model2), distribution="norm", envelope=0.95,
           main="Half-Normal Plot of Residuals", ylab="Residuals")

# Predictions
poly_ck$ck_cut_predicted <- predict(model2, poly_ck)

# Plot predicted vs observed
ggplot(poly_ck, aes(x=ck_cut, y=ck_cut_predicted)) +
  geom_point(alpha=0.5) +
  geom_abline(intercept=0, slope=1, color="red") +
  labs(title="Predicted vs Observed CK", x="Observed", y="Predicted") +
  theme_minimal()
```

2) fit a log trasform

```{r}
model3 <- lm(log(ck_cut) ~ gcs_cat + iss_cat + invasive + sex + age_cat + 
             bleeding + fracture + concussion + brain_edema + brain_compression +
             unconsciousness, data = poly_ck)

summary(model3)
```
compare the three

```{r}
AIC(model, model2, model3)
BIC(model, model2, model3)
```



##########################################################
TODO: does it make sense to add regional trauma indicators maybe when modelling ais? So in this way we know if certain injures location are more related to hhigher ais

So for example we could model 
Model AIS_Head and AIS_thorax. Maybe you need to go back to original df and fit a lmer to each AIS scoring