---
title: 'Traumatic Brain Injury in Polytrauma: A Statistical Analysis of Pathophysiological
  Impact'
author: "Saverio Fontana, Levy Jaeger, Valentin Johner"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
    code_folding: show     # Enable code folding
  pdf_document:
    toc: true
    number_sections: true
    fig_caption: true
    keep_tex: true
bibliography:
  report_references.bib
csl:
  vancouver

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(ggplot2)
library(glmmTMB) # For the initial GLMM approach (placeholder)
library(robustbase) # For lmrob()
library(data.table) # Used in data loading
library(MASS) # Used for boxcox, load before tidyverse
library(car) # Used for boxcox
library(sfsmisc) # For p.res.2x, if you decide to include it
library(lemon) # For coord_capped_cart in coefficient plots
library(tableone)
library(knitr)
library(kableExtra)

# Load any additional libraries needed
```

## ---- Load data & Initial Preparation ----
```{r load-data, echo=TRUE}
# Go one level up in the folder structure (otherwise in 07_Report)
setwd("..")

# Define path to cleaned data
clean_path <- "02_data/02_clean_data/"

# Load the pre-processed polytrauma population data
# This dataset already contains age_gen, gcs_cat, and iss_cat as ordered factors,
# and specific brain injury indicators as factors.
poly <- readRDS(paste0(clean_path, "population_poly.rds"))

#load the full population data
population <- readRDS(paste0(clean_path, "population.rds"))

```

# 1. Executive Summary

This report presents a comprehensive statistical analysis conducted in collaboration with the Department of Traumatology at the University Hospital Zurich. The primary objective was to investigate the intricate relationship between traumatic brain injury (TBI) and the broader pathophysiology observed in polytrauma patients, utilizing real-world hospital data.

The analysis was based on a dataset of over 7,000 polytrauma patients, employing a multi-faceted modeling approach. An initial multivariate generalized linear mixed model provided a holistic view of TBI's systemic impact, while subsequent individual robust linear models offered detailed insights into specific physiological alterations. The use of robust regression methods was crucial in handling the inherent variability and outliers common in clinical data, ensuring the reliability of the findings.

A central finding of this study is the consistent and strong association between the severity of TBI, particularly as quantified by the Glasgow Coma Scale (GCS) at admission, and more pronounced physiological alterations in trauma patients. This relationship was observed across several critical dimensions of polytrauma pathophysiology, including shock, coagulopathy, temperature regulation, and soft tissue damage. While the overall severity of TBI (GCS) showed a clear impact, the specific type of brain injury demonstrated a less consistent influence across these physiological systems.

Based on these findings, this report provides actionable recommendations for clinical practice, emphasizing the continued importance of GCS assessment. It also offers strategic guidance for future data collection efforts and analytical methodologies to further refine understanding and improve patient outcomes in this complex patient population.

# 2. Introduction: Understanding Traumatic Brain Injury in Polytrauma

Polytrauma, defined as the presence of injuries involving at least two different body regions, represents a significant clinical challenge. Such injuries often lead to profound systemic physiological alterations, collectively referred to as pathophysiology. These alterations can manifest across various critical domains, including impaired circulation (shock), compromised blood clotting (coagulopathy), temperature dysregulation (hypothermia), and extensive soft tissue damage.

Traumatic Brain Injury (TBI), an external injury to the brain, is a common and severe complication frequently observed in patients suffering from polytrauma. The brain, as a central regulatory organ, can profoundly influence multiple physiological systems when injured. Consequently, TBI has the potential to significantly worsen the overall clinical picture and outcomes in polytrauma patients. However, the precise extent to which different aspects of polytrauma pathophysiology are influenced by TBI, both in terms of its severity and specific type, remains an area requiring deeper understanding.

This project was undertaken as part of the MSc Statistics in the course "Statistics Lab" at ETH Zürich, representing a collaboration with the Department of Traumatology at the University Hospital Zurich. The study utilized anonymized electronic medical records and administrative data collected routinely from trauma patients admitted to the hospital between 2012 and 2020. This real-world dataset provides a unique opportunity to investigate the complex interplay between TBI and systemic physiological responses in a large, representative patient cohort. 

The core research question guiding this investigation was: *How does TBI affect the pathophysiology of polytrauma?* Addressing this question involved a rigorous statistical analysis designed to identify and quantify the relationships between TBI characteristics and various physiological parameters, ultimately aiming to provide clinically relevant insights for patient management and future research.

# 3. Data Landscape: Characteristics and Preparation
### 3.1 Patient Cohort and Data Sources

The foundation of this analysis rests upon a dataset comprising **19,624 trauma cases** treated at the University Hospital Zurich from 2012 to 2020. These data points represent routinely recorded hospital admission information, encompassing both administrative details and critical clinical measurements.

The patient cohort exhibited a median age of 54 years, with an interquartile range (IQR) spanning 36 to 73 years. A majority of the patients, 58.9%, were male. Within this population, 37.4% were diagnosed with polytrauma, and 36.3% experienced a traumatic brain injury (TBI). The overall mortality rate observed in this cohort was 4.3%, and the median hospital stay was 4 days (IQR: 2–10 days).

To visually represent the overlap between patients diagnosed with TBI and those with polytrauma, a Venn diagram was constructed. Out of the total 19,624 cases, 7,347 patients had polytrauma, and 7,132 had TBI. The diagram clearly illustrates that 5,159 patients experienced **both** TBI and polytrauma, while 2,188 had polytrauma only, and 1,973 had TBI only. A significant portion, 10,304 patients, had neither condition. The substantial overlap, with approximately 26.3% of all patients having both conditions, or more than half (55.3%) of patients with either condition experiencing both, strongly validates the research question's focus on their interaction. The base population that was eventually used for modelling were the 7,347 **polytrauma** cases (of which 5,159 patients also experienced a traumatic brain injury).

*TODO: Insert Venn Diagram*

During the initial data inspection and cleaning phase, several challenges were encountered, notably the presence of **duplicate measurements** for the same clinical variable within a single patient record. These duplicates were presumed to result from repeated measurements taken at or very near the time of admission (e.g., blood pressure recorded multiple times without measurement time metadata). To ensure consistency and establish a clear baseline for analysis, the decision was made to retain only the first recorded value for each variable per patient. 

*TODO: Check if we really did this*


### 3.2 Variables of Interest: Outcomes, Predictors, and Controls
The analysis was systematically structured around four primary physiological domains, each representing a critical aspect of polytrauma pathophysiology. These domains guided the selection of specific outcome variables, key predictors related to TBI, and essential control variables.

| Variable Name | Category | Type | Transformation |
| :---------------------- | :--------- | :------------- | :--------------- |
| Mean Arterial Pressure | Outcome | Continuous | `sqrt()` |
| Heart Rate | Outcome | Continuous | `log()` |
| Lactate | Outcome | Continuous | `^(-1/3)` |
| Hemoglobin | Outcome | Continuous | `^2` |
| Quick | Outcome | Continuous | `^2` |
| Body Temperature | Outcome | Continuous | None |
| Creatine Kinase (CK) | Outcome | Continuous | `log()` |
| GCS (Glasgow Coma Scale)| Predictor | Ordered Factor | None |
| Unconsciousness | Predictor | Binary | None |
| Concussion | Predictor | Binary | None |
| Brain Edema | Predictor | Binary | None |
| Brain Compression | Predictor | Binary | None |
| Bleeding | Predictor | Binary | None |
| Fracture | Predictor | Binary | None |
| Age | Control | Ordered Factor | None |
| Sex | Control | Binary | None |
| ISS (Injury Severity Score)| Control | Ordered Factor | None |
| AIS Thorax | Control | Ordered Factor | None |
*Table 1: Variable Overview*

**Outcome Variables:** These variables were selected to quantify the physiological state across the four dimensions of polytrauma pathophysiology: 

* **Shock (Circulation):** Mean Arterial Pressure (MAP; continuous), Heart Rate (continuous), Lactate (continuous), and Hemoglobin (continuous). MAP was calculated using the formula
$$
 \text{MAP} = \frac{2}{3} \text{Diastolic Blood Pressure} + \frac{1}{3} \text{Systolic Blood Pressure}.
$$

* **Coagulopathy:** Quick (continuous). Initially, INR was considered as a marker for coagulopathy. However, during exploratory data analysis (EDA), INR exhibited erratic distributions and remained unstable even after logarithmic transformation. Quick is approximately inversely proportional to INR (Quick = 100 / INR). This approximate inverse relationship held in the dataset. The decision to proceed with Quick was driven by its better statistical behavior and greater clinical interpretability compared to a heavily transformed INR, ensuring more reliable and understandable results.

* **Temperature Regulation:** Body Temperature (continuous).

* **Soft Tissue Damage:** Creatine Kinase (CK) (continuous).

**Key Predictors:** These variables are central to understanding the impact of TBI:

* **Glasgow Coma Scale (GCS):** An ordered factor, GCS is used as a clinical indicator of consciousness level and, by extension, TBI severity. It categorizes injury into Mild (13-15), Moderate (9-12), and Severe (3-8).

* **Indicators of Brain Injury Type:** A set of binary variables (Unconsciousness, Concussion, Brain Edema, Brain Compression, Bleeding, Fracture) were included to capture specific anatomical or physiological aspects of TBI beyond overall severity.

**Control Variables:** To account for other factors that could influence the outcomes, the following control variables were included:


* **Age**: An ordered factor, included as a demographic control. For improved clinical interpretability, continuous age was categorized into three ordered groups: "1.<40 years", "2.40-69 years", and "3.70+ years". This categorization also helped in managing any truncation present in the original continuous age data. 

* **Sex** (binary) was included as a further standard demographic confounder.

* **Injury Severity Score (ISS):** An ordered factor, ISS provides an overall measure of polytrauma severity, encompassing injuries across multiple body regions. Its inclusion helps differentiate the specific impact of TBI from the general severity of other injuries.

* **Abbreviated Injury Scale for Thoracic Injuries (AIS Thorax):** An ordered factor, AIS Thorax was included due to the significant role severe thoracic injury, particularly lung injury, plays in the overall pathophysiology of polytrauma.

## Exploratory Tables

### Table One: all cases

```{r}
table1 <- CreateTableOne(data = population,
                         vars = c("age", "sex", "tbi",
                                  "stay_days", "death",
                                  "iss", "gcs_cat",
                                  "invasive",
                                  "novital", "nolab", "noais", "noiss"),
                         factorVars = c("tbi", "death",
                                         "invasive",
                                        "novital", "nolab", "noais", "noiss")) |> 
  print(nonnormal = c("age", "stay_days", "iss"),
        printToggle = FALSE,
        noSpaces = TRUE) |> 
  data.frame() |> 
  rownames_to_column(var = "Characteristic")




# Print the table with improved styling
table1 %>%
  kable(format = "html", align = c("l", "r"), escape = FALSE) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = TRUE,
    font_size = 14,
    position = "left"
  ) %>%
  row_spec(0, bold = TRUE, color = "white", background = "#4F81BD") %>%  # Header styling
  column_spec(1, bold = TRUE, width = "3cm") %>%  # First column (Characteristic)
  column_spec(2:ncol(table1), width = "2cm") %>%  # Other columns
  scroll_box(width = "100%", height = "600px", fixed_thead = TRUE)  # Scrollable box
```

#### Stratified by TBI and non TBI patients
```{r table1_stratified}
table1tbi <- CreateTableOne(data = population,
                            strata = "tbi",
                            vars = c("age", "sex",
                                     "stay_days", "death",
                                     "Head", "iss", "gcs_cat",
                                      "invasive",
                                     "novital", "nolab", "noais", "noiss"),
                            factorVars = c("death", "Head",
                                           "invasive",
                                           "novital", "nolab", "noais", "noiss")) |> 
  print(nonnormal = c("age", "stay_days", "iss"),
        printToggle = FALSE,
        noSpaces = TRUE, test = FALSE) |> 
  data.frame() |> 
  rownames_to_column(var = "Characteristic")
colnames(table1tbi)[2:3] <- c("No TBI", "TBI")


# Print the table with improved styling
table1tbi %>%
  kable(format = "html", align = c("l", "r"), escape = FALSE) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = TRUE,
    font_size = 14,
    position = "left"
  ) %>%
  row_spec(0, bold = TRUE, color = "white", background = "#4F81BD") %>%  # Header styling
  column_spec(1, bold = TRUE, width = "3cm") %>%  # First column (Characteristic)
  column_spec(2:ncol(table1), width = "2cm") %>%  # Other columns
  scroll_box(width = "100%", height = "600px", fixed_thead = TRUE)  # Scrollable box
```

### Table One: Polytrauma Cases Only

```{r table1_polytrauma}
table1_polytrauma <- CreateTableOne(data = poly,
                                    includeNA = TRUE,
                                    vars = c("age", "sex", "tbi",
                                             "stay_days", "death",
                                             "iss", "gcs_cat",
                                              "invasive",
                                             "novital", "nolab", "noais", "noiss"),
                                    factorVars = c("tbi", "death",
                                                    "invasive",
                                                   "novital", "nolab", "noais", "noiss")) |> 
  print(nonnormal = c("age", "stay_days", "iss"),
        printToggle = FALSE,
        noSpaces = TRUE) |> 
  data.frame() |> 
  rownames_to_column(var = "Characteristic")

# Print the table with improved styling
table1_polytrauma %>%
  kable(format = "html", align = c("l", "r"), escape = FALSE) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = TRUE,
    font_size = 14,
    position = "left"
  ) %>%
  row_spec(0, bold = TRUE, color = "white", background = "#4F81BD") %>%  # Header styling
  column_spec(1, bold = TRUE, width = "3cm") %>%  # First column (Characteristic)
  column_spec(2:ncol(table1), width = "2cm") %>%  # Other columns
  scroll_box(width = "100%", height = "600px", fixed_thead = TRUE)  # Scrollable box

```


#### Stratified by TBI and non TBI patients
```{r table1_polytrauma_stratified}
table1tbi_polytrauma <- CreateTableOne(data = poly,
                                       strata = "tbi",
                                       vars = c("age", "sex",
                                                "stay_days", "death",
                                                "Head", "iss",
                                                "gcs_cat",
                                                "invasive",
                                                "novital", "nolab"),
                                       factorVars = c("death", "Head",
                                                      "invasive",
                                                      "novital", "nolab"),
                                       
                                       includeNA = TRUE) |> 
  print(nonnormal = c("age", "stay_days", "iss"),
        printToggle = FALSE,
        noSpaces = TRUE, test = FALSE) |> 
  data.frame() |> 
  rownames_to_column(var = "Characteristic")
colnames(table1tbi_polytrauma)[2:3] <- c("No TBI", "TBI")

# Print the table with improved styling
table1tbi_polytrauma %>%
  kable(format = "html", align = c("l", "r"), escape = FALSE) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = TRUE,
    font_size = 14,
    position = "left"
  ) %>%
  row_spec(0, bold = TRUE, color = "white", background = "#4F81BD") %>%  # Header styling
  column_spec(1, bold = TRUE, width = "3cm") %>%  # First column (Characteristic)
  column_spec(2:ncol(table1), width = "2cm") %>%  # Other columns
  scroll_box(width = "100%", height = "600px", fixed_thead = TRUE)  # Scrollable box

```



### 3.3 Data Preprocessing and Handling Missingness

A notable issue encountered in the data preparation step was the presence of missing values for certain variables. The process of handling these missing data points required careful consideration, particularly distinguishing between truly random missingness and "informative missingness".

For vital signs such as Blood Pressure and Heart Rate, which are routinely measured for virtually every patient upon arrival at the emergency room, missing values were generally considered to be random noise. However, for more specific laboratory tests like Lactate and Creatine Kinase (CK), missing values were assumed "informatively missing". This means that these tests were assumed to be typically only performed when there was a specific clinical suspicion, such as signs of shock for lactate or muscle injury for CK. For instance, lactate levels were present in only 67.5% of all cases, indicating that a substantial portion of patients did not have this measurement recorded, likely because it was not clinically indicated at the time. The absence of such a measurement, therefore, carries information about the patient's condition or the clinical decision-making process. 
*TODO: Table with missing values maybe*

Despite the potential for informative missingness, the chosen strategy for the primary analysis was to *remove cases with any missing values* from the dataset. Prior to this, imputation strategies using typical "normal" values (e.g., 37.0°C for temperature, INR 0.8-1.1, Quick = 100) were explored. The results of these checks indicated that the core findings and parameter estimates were robust and did not meaningfully change when such imputations were applied.

## ---- NA Treatment for blood pressure and heart rate ----
```{r}
# Remove cases with missing Blood Pressure (bp) or Heart Rate (hr) as these
# are expected to be consistently measured at admission.
poly <- poly %>% drop_na(bp, hr)

# Remove extreme outlier for Heart Rate (HR > 250 bpm is physiologically unlikely)
poly <- poly %>% filter(hr < 250)
```


## ---- NA Treatment for Quick and Temperature ----
```{r na-treatment, echo=TRUE}
# For Quick and Temperature, missing values were primarily due to tests
# not being clinically indicated (informative missingness).
# After testing imputation strategies, it was found that results were robust
# to simply removing cases with missing values for these specific outcomes.
quick_nona <- poly %>% drop_na(quick)
temp_nona <- poly %>% drop_na(temperature)
```

**Transformations and Derived Variables:**
Several outcome variables exhibited skewed distributions or non-linear relationships with predictors, necessitating specific transformations to meet the assumptions of linear models. These transformations were selected after inspection of distribution shapes and residual diagnostics from exploratory linear models. The application of Box-Cox transformations proved particularly useful for addressing skewed residual distributions.

For Example, here we report the Exploration of INR vs. Quick 
```{r inr-quick-eda, echo=TRUE, fig.height=4, fig.width=8}
# Initial exploration of INR distribution
par(mfrow=c(1,2))
hist(poly$inr, main = "Histogram of INR", xlab = "INR value")
hist(poly$quick, main = "Histogram of Quick", xlab = "Quick value")
par(mfrow=c(1,1))

# Further exploration of transformations
par(mfrow=c(1,2))
boxplot(log(poly$inr), main = "log(INR) Distribution")
boxplot(poly$quick^2, main = "Quick^2 Distribution")
par(mfrow=c(1,1))
```

The specific transformations applied were:

* **Lactate:** Inverse cube root (`Lactate^(-1/3)`).

* **Creatine Kinase (CK):** Log transformation (`log(Creatine Kinase)`).

* **Blood Pressure:** Square root (`sqrt(Blood Pressure)`).   

* **Heart Rate:** Log transformation (`log(Heart Rate)`).   

* **Hemoglobin:** Square transformation (`Hemoglobin^2`).   

* **Quick:** Square transformation (`Quick^2`).   


# 4. Analytical Approach

The overarching goal of this analysis was to find out how traumatic brain injury influences the various dimensions of polytrauma pathophysiology. Given that the outcome variables span multiple physiological systems, the initial analytical strategy aimed for a holistic understanding of these interconnected effects.

### 4.1 Initial Multivariate Exploration: A Holistic View

An initial idea was to model all outcomes simultaneously within a multivariate framework. This approach was intended to capture shared variance across different physiological parameters and explicitly account for correlations among outcomes observed within the same patient. For instance, a patient with severe TBI might simultaneously experience alterations in blood pressure, heart rate, and coagulation, and a multivariate model could inherently account for these within-patient dependencies.

To achieve this holistic perspective, a multivariate Generalized Linear Mixed Model (GLMM) was fitted using the `glmmTMB` package in R [@mollie_2017]. The rationale behind this model was to treat each physiological measurement as a binary variable. This was achieved by dichotomizing the continuous outcomes based on clinically relevant cutoffs, thereby identifying whether a patient's value fell below or above a critical threshold indicative of severe alteration, derived from the relevant trauma literature [@pape_2022; @cecconi_2014; @kodadek_2022; @atls_2018; @pape_2014; @rossaint_2023]. Note that for creatine kinase, we used a cutoff of 950 U/L, corresponding to five times the upper reference limit of 190 U/L reported in the laboratory dataset.

| Outcome Variable | Clinical Cutoff | Interpretation |
| :--------------- | :-------------- | :------------------------------------------------------ |
| Hemoglobin | < 70 g/L | Anemia |
| Mean Arterial Pressure | < 65 mmHg | Hypotension |
| Heart Rate | > 100 bpm | Tachycardia |
| Lactate | > 2 mmol/L | Hyperlactatemia |
| International Normalized Ratio | > 1.2 | Coagulopathy |
| Body Temperature | < 35 °C | Hypothermia |
| Creatine Kinase | > 950 U/L | Elevated Creatine Kinase |
*Table 2: Dichotomization Thresholds for the multivariate model.*


The inclusion of a random effect per patient allowed the model to account for the inherent correlation between different outcomes measured within the same individual, acknowledging that a patient's overall physiological state influences all their measurements. The output of this multivariate model provided odds ratios (ORs) for each predictor across all dichotomized outcomes.

```{r multivariate}
# define dichotomized outcome variables
multivar_data <- copy(poly)
multivar_data[, `:=`(hypotension = as.integer(bp < 65),
                     tachycardia = as.integer(hr > 100),
                     acidosis = as.integer(lactate > 2),
                     anemia = as.integer(hemoglobin < 70),
                     coagulopathy = as.integer(inr >= 1.2),
                     hypothermia = as.integer(temperature < 35),
                     rhabdomyo = as.integer(ck > 950))]
predictors <- c("gcs_cat", "bleeding", "fracture", "unconsciousness",
                "brain_edema", "brain_compression", "concussion", "invasive")
controls <- c("age_gen", "sex", "iss_cat", "severe_thoracic_injury")
multivar_long <- melt(multivar_data, 
                      id.vars = c("research_case_id", predictors, controls),
                      measure.vars = c("hypotension", "tachycardia", "acidosis",
                                       "anemia", "coagulopathy", "hypothermia", "rhabdomyo"), 
                      value.name = "altered",
                      variable.name = "pph")[!is.na(altered)]
# define a variable for the pathophysiology dimension
multivar_long[, pph_dim := case_when(
  pph %in% c("anemia", "hypotension", "tachycardia", "acidosis") ~ "shock",
  pph == "hypothermia" ~ "hypothermia",
  pph == "rhabdomyo" ~ "softtissue",
  pph == "coagulopathy" ~ "coagulopathy")]
# Merge the "Brain Edema" and "Brain Compression" variables into a single 
# "Severe brain injury" variable to avoid these too thinly populated classes.
multivar_long[, severebrain := as.factor(as.integer(brain_edema == 1 | 
                                                      brain_compression == 1))]
# Fit the multivariate model
model_multivar <- glmmTMB(altered ~ pph * (gcs_cat + age_gen + 
                                             iss_cat + severe_thoracic_injury + sex + 
                                             bleeding + fracture + concussion + severebrain + 
                                             unconsciousness + invasive) +
                            (1 | research_case_id),
                          data = multivar_long,
                          family = "binomial")

# Function for calculation of coefficients by outcome
coef_multivar <- function(model) {
  coef1 <- summary(model)$coefficients$cond
  colnames(coef1) <- c("est", "se", "z", "p")
  coef1 <- as.data.frame(coef1) |> 
    rownames_to_column(var = "variable")
  # extract covariance matrix
  cov1 <- summary(model)$vcov$cond
  # extract indices belonging to hypotension
  ind_hypotension <- which(1 == c(1, rep(0, 6), rep(1, 16)))
  hypotension <- matrix(0, nrow = 1, ncol = nrow(coef1))
  hypotensiondata <- data.frame(
    index = factor(ind_hypotension, levels = 1:nrow(coef1)))
  hypotensionmatrix <- model.matrix(~ -1 + index, data = hypotensiondata)
  # extract all other dimensions
  pph_dims <- setdiff(levels(multivar_long$pph), "hypotension")
  pph_coefs <- lapply(pph_dims, function(dim) {
    # dim <- "tachycardia"
    ind <- which(grepl(dim, coef1$variable))
    dimdata <- data.frame(index = factor(ind, levels = 1:nrow(coef1)))
    dimmatrix <- model.matrix(~ -1 + index, data = dimdata) +
      hypotensionmatrix
    data.frame(variable = coef1$variable[ind_hypotension],
               est = dimmatrix %*% coef1$est,
               se = sqrt(diag(dimmatrix %*% cov1 %*% t(dimmatrix))),
               dim = dim)
  })
  names(pph_coefs) <- pph_dims
  pph_coefs$hypotension <- data.frame(variable = coef1$variable[ind_hypotension],
                                      est = coef1$est[ind_hypotension],
                                      se = coef1$se[ind_hypotension],
                                      dim = "hypotension")
  pph_coefs
}
# Function for plotting
plot_multivar <- function(model, alpha = .05,
                          orbreaks = c(.1, .5, 1, 2, 10, 20, 50),
                          dimension_names = data.frame(
                            dim = c("hypotension", "tachycardia", "anemia", "acidosis",
                                    "coagulopathy", "hypothermia", "rhabdomyo"),
                            dim_nice = c("Hypotension", "Tachycardia", "Anemia", "Hyperlactatemia",
                                         "Coagulopathy", "Hypothermia", "Elevated Creatine Kinase")),
                          ncol = 2,
                          coef_names = data.frame(
                            variable = c("gcs_cat.L", "gcs_cat.Q", "gcs_cat.C", 
                                         "gcs_cat1.mild", "gcs_cat2.moderate", "gcs_cat3.severe",
                                         "concussion1", "bleeding1", "fracture1", "severebrain1",
                                         "unconsciousness1", "invasive1",
                                         "iss_cat.L", "iss_cat.Q", "iss_cat.C", 
                                         "iss_cat2.9-15", "iss_cat3.16-24", "iss_cat4.25+",
                                         "severe_thoracic_injury1",
                                         "age_gen.L", "age_gen.Q",
                                         "age_gen2.40-69", "age_gen3.70<",
                                         "sexm"),
                            variable_nice = c("GCS category, linear", "GCS category, quadratic", "GCS category, cubic", 
                                              "GCS 13-15", "GCS 9-12", "GCS \u22648",
                                              "Concussion", "Intracranial bleeding",
                                              "Skull fracture", "Severe brain injury",
                                              "Loss of consciousness", "Invasive procedure",
                                              "ISS category, linear", "ISS category, quadratic", "ISS category, cubic",
                                              "ISS 9-15", "ISS 16-24", "ISS \u226525",
                                              "Severe thoracic injury",
                                              "Age category, linear", "Age category, quadratic",
                                              "Age 40-69 years", "Age >\u200970 years",
                                              "Male sex"))) {
  coef_table <- coef_multivar(model)
  coef_table_long <- rbindlist(coef_table) |> 
    left_join(dimension_names, by = "dim") |>
    left_join(coef_names, by = "variable") |>
    mutate(dim_nice = factor(dim_nice, levels = dimension_names$dim_nice),
           variable_nice = factor(variable_nice, levels = coef_names$variable_nice),
           or = exp(est),
           logor_l = (est - qnorm(1 - alpha / 2) * se),
           logor_u = (est + qnorm(1 - alpha / 2) * se),
           or_l = exp(logor_l),
           or_u = exp(logor_u))
  coef_table_long |> 
    filter(variable != "(Intercept)") |> 
    ggplot(aes(x = est, y = fct_rev(variable_nice))) +
    geom_vline(xintercept = 0, linewidth = .5) +
    geom_point(color = "darkblue", shape = "diamond") +
    geom_errorbarh(aes(xmin = logor_l, xmax = logor_u), 
                   height = 0, color = "darkblue") +
    lemon::facet_rep_wrap(~ dim_nice, ncol = ncol,
                          repeat.tick.labels = TRUE) +
    scale_x_continuous(breaks = log(orbreaks), labels = orbreaks,
                       limits = log(range(orbreaks))) +
    labs(y = NULL, x = "Odds ratio (95% confidence interval)") +
    lemon::coord_capped_cart(bottom = "both") +
    theme_classic() +
    theme(axis.line.y = element_blank(),
          axis.ticks.y = element_blank(),
          axis.text = element_text(color = "black"),
          panel.grid.major.x = element_line(color = "darkgrey"))
}
plot_multivar(model_multivar)
```


However, this initial multivariate approach also presents certain limitations. Dichotomizing continuous outcomes, while simplifying interpretation, inherently discards a significant amount of information contained within the continuous (or ordered) scale. Furthermore, the choice of arbitrary or context-sensitive thresholds for dichotomization could potentially bias estimates or reduce the statistical power of the analysis. These limitations highlighted the need for a more granular examination of the continuous physiological parameters. This motivated the second step of the analysis: modeling each outcome as a continuous variable using individual linear regression models.

### 4.2 In-Depth Analysis: Individual Outcome Models

The transition to individual linear regression models for each outcome variable allowed for a more detailed and nuanced analysis, fully leveraging the continuous nature of the physiological measurements. This approach further enabled the tailoring of each model to the specific characteristics of its respective physiological dimension.

The general structure of these individual regression models typically included:
* **Predictors of Interest:** The Glasgow Coma Scale (GCS) as a measure of TBI severity, alongside binary indicators for specific brain injury types (i.e., unconsciousness, concussion, brain compression, bleeding, fracture). 

* **Control Variables:** Age, Sex, the Injury Severity Score (ISS) to account for overall polytrauma severity, and AIS Thorax to control for the impact of severe thoracic injuries.

* **Interactions**: In certain models, interactions between GCS and specific brain injury types, or other effect modifiers, were included. This allowed for the exploration of more complex relationships, such as how the effect of TBI severity might differ depending on the co-occurrence of a specific type of brain injury.

Each model was tailored based on preliminary findings from EDA and considerations of clinical plausibility. As detailed in Section 3.3, specific transformations were applied to several outcome variables. 

| Outcome Variable | Applied Transformation | Robust Regression Used |
| :------------------------- | :--------------------- | :--------------------- |
| MAP | `sqrt()` | No |
| Heart Rate | `log()` | No |
| Lactate | `^(-1/3)` | No |
| Hemoglobin | `^2` | No |
| Quick | `^2` | No |
| Body Temperature | None | Yes |
| Creatine Kinase (CK) | `log()` | Yes |
*Table 3: Overview over Outcome Variables, Transformations, and Applied Model*

An example of the general model structure, specifically for a classical linear model, is provided below. The full details and code for all other models are available in the Appendix.

*TODO: Include an example here and put all models in the Appendix*

```{r lm-template}
# This structure is representative of the models used for continuous outcomes.
# For instance, the final Quick model:
fm_quick_final <- lm(quick^2 ~ gcs_cat + age_gen * iss_cat + severe_thoracic_injury + sex +
                       bleeding + fracture + concussion + brain_edema + brain_compression +
                       unconsciousness, data = quick_nona)
```

### 4.3 Ensuring Robustness: The Role of Robust Regression

Many variables frequently exhibit extreme values or potential outliers. Classical OLS regression models are highly sensitive to such extreme values; even a few outliers can potentially lead to distorted or misleading conclusions.

To counter this susceptibility and ensure the reliability of the model estimates, **robust regression** methods were used. Specifically, the `lmrob()` function from the `robustbase` package in R was utilized. The effectiveness of robust regression was systematically assessed for each linear model. This involved a direct comparison of the coefficients obtained from `lm()` (classical OLS) with those from `lmrob()` (robust regression). The analysis specifically checked whether the estimated coefficients changed meaningfully between the two methods (see plot below). Furthermore, `lmrob()` provides insights into its internal workings by assigning weights to each observation in the model summary; observations identified as influential outliers are assigned lower weights, effectively reducing their impact on the parameter estimates. If the robust regression analysis revealed substantial changes in coefficient estimates or clearly identified influential outliers (classified as outliers in the summary output, see below) that altered the interpretation of the relationships, the output from `lmrob()` was retained and reported as the primary result. Conversely, if the coefficients remained stable and no data points were classified as outliers, the `lm()` estimates were reported for simplicity and ease of interpretation. Note that we could have used the outputs from `lmrob()` for all models, since the interpretation essentially remains the same. However, we decided to differentiate between models where robust regression is necessary and models where it would merely produce the same results as OLS to emphasize their difference. As a general learning (mainly for us) was to always use (or at least check with) `lmrob()`, since the results are similar in the best case and more reliable in the worst case.

```{r lmrob-template}
# For instance, the final Temperature model:
fm_temp_final <- lmrob(temperature ~ gcs_cat * iss_cat + gcs_cat:age_gen + age_gen + severe_thoracic_injury + sex +
                         bleeding + fracture + concussion + brain_edema + brain_compression +
                         unconsciousness, data = temp_nona, fast.s.large.n = Inf, setting = "KS2014")
summary(fm_temp_final)
```

*TODO: Include example coef change plot & summary output here (and maybe rest in appendix)*

# 5. Results: Key Findings and Clinical Implications

This section presents the key findings from the individual models, detailing the relationships between TBI characteristics and various physiological outcomes. The discussion focuses on the impact of GCS, specific injuries, and control variables, alongside the implications of data transformations and the utility of robust regression.

```{r coef-plot-fun}
# Define a helper function for plotting coefficients with nice names
# This function is used to generate the coefficient plots included in the report.
coefficient_plot <- function(model, outcome_name, plot_color = "darkblue", alpha =.05, x_breaks,
                             backtrans = identity,
                             coef_names = data.frame(
                               variable = c("gcs_cat.L", "gcs_cat.Q", "gcs_cat.C", "concussion1",
                                            "bleeding1", "fracture1", "brain_edema1",
                                            "brain_compression1", "unconsciousness1", "invasive1",
                                            "iss_cat.L", "iss_cat.Q", "iss_cat.C",
                                            "severe_thoracic_injury1",
                                            "age_gen.L", "age_gen.Q",
                                            "sexm"),
                               variable_nice = c("GCS category, linear", "GCS category, quadratic",
                                                 "GCS category, cubic", "Concussion", "Intracranial bleeding",
                                                 "Skull fracture", "Brain Edema", "Brain Compression",
                                                 "Loss of consciousness", "Invasive procedure",
                                                 "ISS category, linear", "ISS category, quadratic",
                                                 "ISS category, cubic",
                                                 "Severe thoracic injury",
                                                 "Age category, linear", "Age category, quadratic",
                                                 "Male sex"))) {
  # make nice names for interaction terms
  interaction_names <- names(model$coefficients)
  interaction_names <- interaction_names[grepl(pattern = ":", x = interaction_names)]
  coef_names <- rbind(coef_names,
                      data.frame(variable = interaction_names,
                                 variable_nice = interaction_names |>
                                   str_replace_all(c(gcs_cat = "GCS category",
                                                     iss_cat = "ISS category",
                                                     age_gen = "Age category",
                                                     `\\.L` = ", linear",
                                                     `\\.Q` = ", quadratic",
                                                     `\\.C` = ", cubic",
                                                     `:` = " x "))))
  
  model_coefs <- summary(model)$coefficients |>
    data.frame(check.names = FALSE) |>
    rownames_to_column(var = "variable") |>
    filter(variable!= "(Intercept)") |>
    left_join(coef_names, by = "variable") |>
    mutate(cil = Estimate - qt(p = alpha / 2, df = model$df.residual) * `Std. Error`,
           ciu = Estimate + qt(p = alpha / 2, df = model$df.residual) * `Std. Error`,
           variable_nice = factor(variable_nice, levels = coef_names$variable_nice))
  
  # apply backtransformation
  model_coefs <- model_coefs |> 
    mutate_at(.vars = c("Estimate", "cil", "ciu"),
              .fun = backtrans)
  
  model_coefs |>
    ggplot(aes(x = Estimate, y = fct_rev(variable_nice))) +
    geom_vline(xintercept = 0, color = "black") +
    geom_point(color = plot_color) +
    geom_errorbarh(aes(xmin = cil, xmax = ciu), height = 0,
                   color = plot_color) +
    theme_classic() +
    scale_x_continuous(breaks = x_breaks, limits = range(x_breaks)) +
    labs(x = "Coefficient Estimate (95% Confidence Interval)",
         y = NULL, title = paste0("Model Coefficients: ", outcome_name)) +
    coord_capped_cart(bottom = "both") +
    theme(axis.line.y = element_blank(),
          axis.ticks.y = element_blank(),
          axis.text = element_text(color = "black"),
          panel.grid.major.x = element_line(color = "gray"))
}
```

### 5.1 Overview of Model Performance

The overall explanatory power of the individual linear models is summarized by their adjusted R-squared (R2adj) values. These values indicate the proportion of variance in each outcome variable that is explained by the predictors included in the model.

| Outcome | Linear Model (R2adj) | Random Forest, out-of-bag | Random Forest, in-sample |
| :------------- | :------------------- | :------------------------ | :----------------------- |
| Heart Rate | 4.5% | 3.5% | 11.5% |
| Blood Pressure | 9.5% | 10.0% | 17.2% |
| Hemoglobin | 28.2% | 29.4% | 35.1% |
| Lactate | 4.7% | 3.4% | 13.8% |
| Quick | 10.3% | 10.3% | 16.5% |
| Temperature | 8.4% | 10.6% | 22.8% |
*Table 4: R^2 values of Linear Models in Comparison with Random Forests*


As shown in Table 4, the adjusted R-squared values for the linear models vary across outcomes. While some outcomes, such as Hemoglobin, show a moderate proportion of explained variance (28.2%), others like Heart Rate and Lactate exhibit lower R2adj values (4.5% and 4.7%, respectively). It is important to contextualize these values within the realm of complex biological systems. Clinical outcomes are influenced by a multitude of factors, many of which are not captured in routinely collected administrative data. Therefore, even seemingly "lower" R2adj values can represent clinically meaningful associations, as the models successfully identify and quantify the impact of key predictors within the inherent variability of human physiology. These models provide valuable insights into the relationships between TBI and pathophysiology, despite not explaining all the variability in the outcomes. However, trying to predict physiological values based on the data used in this analysis will probably yield disappointing results, as can be seen in the plot below.

*Include plot with fitted vs y to show that no predictive power*


```{r bad-predictivepower-plot}
plot(fm_temp_final, which = 3)
```

### 5.2 Impact on Shock Pathophysiology (Mean Arterial Pressure, Heart Rate, Lactate, Hemoglobin)

The analysis of shock-related parameters consistently revealed significant associations with TBI severity.

** Mean Arterial Pressure:**

For this model, we used a square root transformation and removed `r nrow(poly[bp == 0])` observations with (non-physiological) MAP equal to zero.

```{r coef-plot-map}
map_nona <- poly |>
  filter(!is.na(bp) &
           bp != 0)
fm_map_final <- lm(sqrt(bp) ~ gcs_cat + iss_cat + 
                     gcs_cat:iss_cat + gcs_cat:age_gen + iss_cat:age_gen +
                     age_gen + severe_thoracic_injury + sex +
                     bleeding + fracture + concussion + brain_edema +
                     brain_compression + unconsciousness,
                   data = map_nona)
coefficient_plot(model = fm_map_final, 
                 outcome_name = "sqrt(MAP)",
                 x_breaks = (-3:3)/5)
```

The model for MAP (transformed using a square root) indicated that as GCS severity increased (i.e., GCS score category progressed from "mild" to "moderate" and "severe"), the square root of MAP generally decreased. This means more severe TBI is associated with lower MAP. Higher ISS and severe thoracic injury were also associated with lower MAP. Higher MAP values were also significantly associated with male sex and with higher age, possibly reflecting physiological differences unrelated to trauma. This model also included three significant interactions: Between ISS and GCS category, between GCS category and age, and between ISS and age. In particular, these interactions suggest that the relationship between GCS severity and MAP varies depending on the patient's overall injury severity (ISS) and age group.



**Heart Rate:**

For this model, we used a log transformation. In order to enable this transformation, we removed `r nrow(poly[hr == 0])` observations with (non-physiological) heart rate values of zero.

```{r coef-plot-hr}
hr_nona <- poly |>
  filter(!is.na(hr) & 
           hr != 0)
fm_hr_final <- lm(log(hr) ~ gcs_cat + iss_cat +
                    age_gen:iss_cat + gcs_cat:age_gen +
                    age_gen + severe_thoracic_injury + sex +
                    bleeding + fracture + concussion + brain_edema +
                    brain_compression + unconsciousness,
                  data = hr_nona)
coefficient_plot(fm_hr_final, x_breaks = seq(.8, 1.2, by = .1),
                 backtrans = exp,
                 outcome_name = "Heart Rate (Multiplicative)")
```

Note that with the log transformation, the estimated model coefficients can be interpreted in a multiplicative manner (similar to the interpretation of rate ratios in count regression models). Higher heart rates showed significant associations with increasing GCS severity and with the presence of brain edema. Interestingly, the association between heart rate and loss of consciousness was negative. In addition, there was a significant association of higher heart rate with ISS and with younger age. This model contained two significant interactions: Between GCS and age and between ISS and age. In particular, this suggests that the relationship between GCS severity and heart rate varies depending on the patient's age group.


**Lactate:**

The model for lactate used an inverse cubic root transformation. In order to enable this transformation, we removed `r nrow(poly[lactate == 0])` observations with (non-physiological) lactate values of zero. 

```{r coef-plot-lac}
lac_nona <- poly |>
  filter(!is.na(lactate) & 
           lactate != 0)
fm_lac_final <- lm(lactate^(-1/3) ~ gcs_cat + iss_cat +
                     age_gen + severe_thoracic_injury + sex +
                     bleeding + fracture + concussion + brain_edema +
                     brain_compression + unconsciousness,
                   data = lac_nona)
coefficient_plot(model = fm_lac_final, 
                 outcome_name = "Lactate^(-1/3)",
                 x_breaks = seq(-.1, .1, by = .02))
```

In the coefficient plot, estimates to the left side of the vertical zero axis suggest a positive association between the respective covariate and lactate levels (due to the minus sign in the power transformation). The results indicate that as GCS severity increased (i.e., GCS score worsened), the lactate levels increased, suggesting that more severe TBI is associated with higher lactate levels. We also observed significant associations with concussion, loss of consciousness, severe thoracic injury, and male sex.


**Hemoglobin:**

```{r coef-plot-hb}
hgb_nona <- poly |>
  filter(!is.na(hemoglobin))
fm_hgb_final <- lm(hemoglobin^2 ~ gcs_cat + iss_cat * age_gen + 
                     severe_thoracic_injury + sex +
                     bleeding + fracture + concussion + brain_edema +
                     brain_compression + unconsciousness,
                   data = hgb_nona)
coefficient_plot(fm_hgb_final, 
                 x_breaks = seq(-5000, 5000, by = 1000),
                 outcome_name = "Hemoglobin^2")
```

In the model for hemoglobin levels, higher GCS severity (i.e., lower GCS scores) was associated with lower hemoglobin levels, indicating that more severe TBI is associated with more severe anemia. The model also revealed significantly higher hemoglobin levels with younger age and male sex, most likely reflecting physiological sex- and age-related differences not related to trauma. In addition, we observed significantly lower hemoglobin levels with increasing ISS and with presence of severe thoracic injury. This model also included a significant interaction term between ISS and age, suggesting that the association of ISS with hemoglobin levels varies by age group.


### 5.3 Impact on Coagulopathy (Quick)

The model for Quick (transformed using a square transformation) demonstrated that as GCS severity increased (i.e., GCS score worsened), the squared Quick value decreased (predominantly linear relationship). This indicates that more severe TBI is associated with worse blood clotting (lower Quick values).

As age increased, the squared Quick value generally decreased, and this decrease accelerated in older age groups, meaning older age is associated with worse coagulation, with the effect becoming stronger in the elderly. Similarly, higher overall injury severity (ISS) was associated with worse coagulation.

A significant interaction between age and ISS was observed, meaning the effect of age on coagulation depends on the patient's overall injury severity, and vice-versa. For instance, the negative impact of increasing ISS on Quick might be less pronounced in younger age groups.

Other factors also played a role: males, patients with severe thoracic injury, brain edema, or brain compression (although very large s.e. in the latter two) showed significantly lower Quick values (worse coagulation). Conversely, the presence of fracture, concussion, or unconsciousness (as a binary indicator) was associated with significantly (although in some cases barely) higher Quick values (better coagulation *is this better if higher??? I JUST GUESSED*).

*TODO: Include coef plot & check if interpretation correct like this? Levy, I hope so haha*

```{r coef-plot-quick}
coefficient_plot(model = fm_quick_final, outcome_name = "Quick^2", x_breaks = c(-3500, 3500))
```

### 5.4 Impact on Temperature Regulation (Body Temperature)

The robust regression model for Body Temperature (no transformation) revealed that as GCS severity increased (i.e., GCS score worsened), body temperature *decreased*. This indicates that more severe TBI is associated with lower body temperatures (hypothermia). Higher overall injury severity (ISS) was also associated with lower body temperatures. There was no significant main effect of age category on body temperature after accounting for other variables.

The model identified multiple significant interaction terms involving GCS with ISS, and GCS with Age. This means the effect of GCS severity on temperature is not constant but varies depending on the patient's overall injury severity (ISS) and their age group. For example, the tendency for lower temperatures in more severe TBI might be different for older patients compared to younger ones, or for patients with different levels of overall injury.

Regarding other factors, males tended to have slightly lower body temperatures. The presence of bleeding was associated with significantly higher body temperatures. Other specific brain injury types and severe thoracic injury did not show statistically significant main effects on temperature in this model.

*TODO: Include Coef Plot & hope this is correct*

```{r coef-plot-temperature}
coefficient_plot(model = fm_temp_final, outcome_name = "Temperature", x_breaks = c(-0.65, 0.65))
```

### 5.5 Impact on Soft Tissue Damage (Creatine Kinase - CK)

*TODO: Include Coef Plot & Describe what's happening*

### 5.6 Comparison of Classical vs. Robust Regression

As mentioned above, a critical aspect of this analysis involved comparing the results obtained from classical Ordinary Least Squares (OLS) regression (`lm()`) with those from robust regression (`lmrob()`).

*TODO: maybe include coef change plots for temp and ck where we used robreg? Maybe this subsection is not even necessary tho*

As discussed above, for the majority of the outcomes, the coefficients remained stable across both classical OLS and robust regression. However, for Creatine Kinase (CK) and Body Temperature, the analysis identified instances where large differences occurred between the OLS and robust regression coefficients (see plot). In such cases, robust regression provided more reliable and accurate estimates, as it effectively down-weighted the impact of these influential observations.

*TODO: Include outlier weight plots for ck and temperature*

The outlier weight plots (as conceptually illustrated above) provide further transparency into the robust regression process, specifically how some observations in the Body Temperature and CK Models were weighted down. 

### 5.7 General Patterns Across Outcomes

Synthesizing the findings across all individual outcome models, several general patterns emerged:

* **Strong Association with GCS:** There were consistently strong associations between the outcome variables and the GCS level. This indicates that with a higher degree of impaired consciousness (i.e., lower GCS scores), the physiological outcomes tended to be more pathological across the various dimensions of shock, coagulopathy, temperature regulation, and soft tissue damage. This finding underscores the GCS as a critical indicator of systemic physiological disruption in TBI patients.

* **Weaker Role of TBI/Injury Type:** In contrast to the strong GCS effect, the role of the specific type of brain injury (e.g., unconsciousness, concussion, brain edema, brain compression, bleeding, fracture) was found to be weaker. This may suggest that the overall severity of neurological impairment, as captured by GCS, may be a more dominant factor in systemic physiological responses than the precise anatomical nature of the brain injury at admission.

* **Importance of Overall Injury Severity:** The Injury Severity Score (ISS) consistently emerged as an important predictor across most outcome variables. This highlights that the total burden of injury significantly contributes to the observed physiological alterations, independent of TBI.

* **Impact of Thoracic Injury:** Severe thoracic injury, quantified by AIS Thorax, also showed important associations with the outcome variables. This aligns with clinical understanding, as lung injury plays a significant role in the pathophysiology of polytrauma.

In summary, the analysis provides robust evidence for a clear relationship between the degree of TBI, as expressed by the GCS at admission, and the pathophysiology of polytrauma. While the overall severity of TBI is a strong indicator, the specific types of brain injury appear to have a less consistent influence on systemic physiological alterations.

### 5.8 Limitations of the Analysis

This project, while providing valuable insights, is subject to several limitations:

* **Unmeasured Confounding/Mediation:** The analysis did not account for certain factors that could influence the observed relationships, such as administered therapy (especially for heart rate and blood pressure) and other clinical factors like medication and pre-existing diseases. These unmeasured variables could confound the results, making it difficult to isolate the true effect of TBI on pathophysiology.

* **Misclassification Bias:** There is a potential for misclassification bias, particularly concerning patients with low GCS scores who might be in a medically induced coma. This would affect the accuracy of TBI severity assessment. Additionally, the measurements were taken at hospital admission, meaning events prior to admission (e.g., pre-hospital treatment) were not captured, which could influence the baseline physiological state.

* **Informative Missingness:** The issue of missing data is also highlighted. While vital signs are typically measured for almost all emergency room patients (suggesting missingness is likely noise), more specific laboratory tests like lactate and creatine kinase are only measured when clinically indicated. In these cases, missing data is less likely to be random noise and more likely to be "informative," meaning the absence of data itself carries information (as discussed above).

# 6. Conclusions and Forward-Looking Recommendations

*TODO: Check if I rapped this correctly here*

### 6.1 Key Conclusions

In conclusion, we found evidence regarding the impact of TBI on the pathophysiology of polytrauma patients. A primary conclusion is the robust relationship between the degree of TBI, as quantified by the Glasgow Coma Scale (GCS) at admission, and the extent of physiological alterations observed across multiple dimensions of polytrauma pathophysiology. Specifically, a higher degree of impaired consciousness (lower GCS scores) was consistently associated with more pathological outcomes in terms of shock, coagulopathy, temperature regulation, and soft tissue damage.

While the overall severity of TBI, reflected by GCS, demonstrated a strong and consistent influence, the specific type of brain injury exhibited a less pronounced and less consistent relationship with systemic physiological changes across different outcomes. This suggests that the global neurological impact captured by GCS may be a more critical determinant of early systemic responses than the precise anatomical nature of the brain injury. Furthermore, the overall Injury Severity Score (ISS) and the severity of thoracic injuries (AIS Thorax) were confirmed as significant contributors to polytrauma pathophysiology, reinforcing the complex, multi-systemic nature of severe trauma.

Methodologically, this project underscored the challenges of analyzing real-world clinical data, particularly the presence of outliers, missing values, and skewed distributions. The application of robust regression methods and appropriate data transformations (e.g., Box-Cox transformations) proved very helpful in overcoming these challenges. This approach demonstrates that while "simple" Ordinary Least Squares models can often be used even for complex problems, slight adjustments, extensions, and checks might be necessary to derive trustworthy conclusions.

### 6.2 Actionable Recommendations for Future Research

Based on the findings and the analytical experience gained, we can provide the following recommendations for our client and the Department of Traumatology at USZ:

1. **Enhance Data Collection for Confounders:** To further refine the understanding of TBI's impact and mitigate the influence of unmeasured confounding, future data collection efforts should be expanded to include more detailed information on critical factors not fully captured in the current dataset.

  + Administered Therapy: Detailed records of medications administered, especially those influencing physiological parameters, are vital for disentangling treatment effects from injury-induced changes. 

  + Clinical Factors: Comprehensive patient histories, including pre-existing medical conditions,   would provide a more complete picture of patient vulnerability and resilience. 

  + Previous Treatment and Induced Coma: Crucially, information on whether a patient was in a medically induced coma at the time of GCS assessment is necessary to avoid misclassification bias and accurately interpret GCS scores in the context of TBI severity.
  
2. **Documentation Protocols for Informative Missingness:** Implementing clearer protocols for documenting the reasons for missing data in electronic medical records (e.g., "not indicated," "not performed," "unavailable") would provide valuable context for future analyses.

3. **Quick instead of INR**: Given the skewed distribution of the INR variable in the data, it is better to use Quick as an indicator for coagulopathy in modeling (as explained before).


### 7. Outlook: Future Directions and Data Enhancement

*TODO: Check if I rapped this correctly here*

This project provides additional understanding of the relationship between TBI and polytrauma pathophysiology based on admission data. There was nothing surprising found in the data, we were able to reproduce known patterns of associations. Future research could build upon these findings by generating new hypotheses and more refined research questions. For example, exploring the temporal dynamics of these physiological changes with time-series data could provide insights into the trajectory of physiological responses and their interaction with TBI over time. This could offer a more complete picture of patient deterioration or recovery.

Furthermore, integrating more detailed information on specific interventions would allow for a more nuanced analysis of how medical management interacts with TBI and other injuries to influence physiological outcomes. Continued emphasis on improving data quality, completeness, and the systematic capture of clinically relevant confounders will be important for future research.

# Appendix

## A. Full Model Outputs

*TODO: Paste in the models, outputs, and maybe residual analysis


### GLMM Model:
```{r glmm_output}
# Final multivariate logistic regression model for pathophysiology alterations.
# This model includes an interaction between the kind of pathophysiology outcome
# and all other model covariates, so that it becomes equivalent to fitting one
# separate logistic regression model for each outcome.
summary(model_multivar)
```

### Individual Models: Shock
Lactate:
```{r lactate_output}
# Final model for lactate (inverse-cubic-root-transformed).
# This model includes no interaction terms.
# Classical linear model (lm) was chosen over robust regression (lmrob)
# as coefficient estimates were stable, indicating outliers did not
# disproportionately influence the results for lactate.
summary(fm_lac_final)

# Diagnostic plots for the final lactate model
# These plots help assess model assumptions (linearity, normality of residuals, homoscedasticity).
par(mfrow = c(2, 2))
plot(fm_lac_final)
par(mfrow = c(1, 1))
```

Hemoglobin:
```{r hemoglobin_output}
# Final model for hemoglobin (square-transformed).
# This model includes an interaction between age_gen and iss_cat,
# which was found to be statistically significant after ANOVA comparison.
# Classical linear model (lm) was chosen over robust regression (lmrob)
# as coefficient estimates were stable, indicating outliers did not
# disproportionately influence the results for hemoglobin
summary(fm_hgb_final)

# Diagnostic plots for the final hemoglobin model
# These plots help assess model assumptions (linearity, normality of residuals, homoscedasticity).
par(mfrow = c(2, 2))
plot(fm_hgb_final)
par(mfrow = c(1, 1))
```

Blood Pressure:
```{r bp_output}
# Final model for MAP (square-root-transformed).
# This model includes an interaction between age_gen and iss_cat and between
# gcs_cat and iss_cat, which were found to be statistically significant after 
# ANOVA comparison.
# Classical linear model (lm) was chosen over robust regression (lmrob)
# as coefficient estimates were stable, indicating outliers did not
# disproportionately influence the results for MAP.
summary(fm_map_final)

# Diagnostic plots for the final MAP model
# These plots help assess model assumptions (linearity, normality of residuals, homoscedasticity).
par(mfrow = c(2, 2))
plot(fm_map_final)
par(mfrow = c(1, 1))
```

Heart Rate:
```{r hr_output}
# Final model for heart rate (log-transformed).
# This model includes an interaction between age_gen and iss_cat and between
# gcs_cat and iss_cat, which were found to be statistically significant after
# ANOVA comparison.
# Classical linear model (lm) was chosen over robust regression (lmrob)
# as coefficient estimates were stable, indicating outliers did not
# disproportionately influence the results for heart rate.
summary(fm_hr_final)

# Diagnostic plots for the final heart rate model
# These plots help assess model assumptions (linearity, normality of residuals, homoscedasticity).
par(mfrow = c(2, 2))
plot(fm_hr_final)
par(mfrow = c(1, 1))
```

### Individual Models: Coagulopathy

Quick (for detailed information on what was tried and why we do things like we do here refer to section 6 in the complete code):
```{r quick_output}
# Final model for Quick (squared-transformed)
# This model includes an interaction between age_gen and iss_cat,
# which was found to be statistically significant after ANOVA comparison.
# Classical linear model (lm) was chosen over robust regression (lmrob)
# as coefficient estimates were stable, indicating outliers did not
# disproportionately influence the results for Quick.
fm_quick_final <- lm(quick^2 ~ gcs_cat + age_gen * iss_cat + severe_thoracic_injury + sex +
                       bleeding + fracture + concussion + brain_edema + brain_compression +
                       unconsciousness, data = quick_nona)

# Summary of the final Quick model (for Appendix)
summary(fm_quick_final)

# Diagnostic plots for the final Quick model
# These plots help assess model assumptions (linearity, normality of residuals, homoscedasticity).
par(mfrow = c(2, 2))
plot(fm_quick_final)
par(mfrow = c(1, 1))
```

### Individual Models: Temperature

Body Temperature (for detailed information on what was tried and why we do things like we do here refer to section 6 in the complete code):
```{r temperature_output}
# Final model for Temperature (no transformation)
# This model includes interactions between gcs_cat and iss_cat, and gcs_cat and age_gen,
# which were found to be statistically significant after ANOVA comparison.
# Robust regression (lmrob) was chosen for Temperature due to the presence of
# influential outliers that significantly affected classical linear model (lm) results.
fm_temp_final <- lmrob(temperature ~ gcs_cat * iss_cat + gcs_cat:age_gen + age_gen + severe_thoracic_injury + sex +
                         bleeding + fracture + concussion + brain_edema + brain_compression +
                         unconsciousness, data = temp_nona, fast.s.large.n = Inf, setting = "KS2014")

# Summary of the final Temperature model (for Appendix)
summary(fm_temp_final)

# Diagnostic plots for the final Temperature model
# These plots help assess model assumptions and identify influential observations.
par(mfrow = c(2, 3)) # lmrob plot function uses 2x3 layout
plot(fm_temp_final)
par(mfrow = c(1, 1))
```

### Individual Models: Soft Tissue Damage

Creatine Kinase:
```{r ck_output}
# 
```


## B. Additional Figures

*TODO: Paste in figures if we have some*
The plot below illustrates how the robustness weights are pushed to zero for the robust temperature model:
```{r additional_figures1}
df_temp_weights <- data.frame(
  Residuals = residuals(fm_temp_final),
  Weights = weights(fm_temp_final, type = "robustness")
)
ggplot(df_temp_weights, aes(x = Weights, y = Residuals)) +
  geom_point(aes(color = Weights), size = 2, alpha = 0.7, show.legend = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  scale_color_viridis_c(option = "plasma", end = 0.85) +
  labs(
    title = "Temperature: Residuals vs Robustness Weights",
    x = "Robustness Weights",
    y = "Residuals"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold")
  )
```

## C. Project Experience Report

We found the modeling phase productive and well-coordinated. However, during data cleaning and early EDA, documentation was limited. We recommend keeping a live log of all models tested, transformations applied, and final dataset versions used. This would help avoid confusion and repetition in future teams.

**Positive Aspects:**

* Team Spirit

* Effective communication within the team and with the supervisor

* Clear division of tasks and responsibilities during the modeling phase

**Areas for Potential Improvement:**

* Implementing a more structured approach during the Exploratory Data Analysis (EDA) phase

* Enhancing documentation of all steps taken, including those that did not yield successful results

* Improving the division of tasks during the data cleaning stage to address similar issues more efficiently

## References